---
title: "Sleep Deprivation and Cognitive Performance"
format: html
editor: visual
---

## Sleep Deprivation and Cognitive Performance

## Goals

-   Understand the effects of sleep deprivation (hours of sleep, sleep quality, daytime sleep) and behavioral determinants of health (BMI, caffeine intake, physical activity level) on cognitive performance (attention, working memory, reaction speed, emotion regulation).
-   Model each cognitive performance metric using sleep and behavioral health features.

## Read Data

```{r read_data}
library(readr)

df <- read_csv("sleep_deprivation_dataset_detailed.csv")
head(df)
```

## Exploratory Data Analysis (EDA)

### Data Structure

```{r data_structure}
str(df)
```

### Summary Statistics

```{r summary}
summary(df)
```

```{r unique_gender}
cat("Unique values of df$Gender:", unique(df$Gender), "\n")
cat("Unique values of df$Caffeine_Intake", sort(unique(df$Caffeine_Intake)), "\n")
```

Note:

-   `Caffeine_Intake`, `Sleep_Quality_Score`, `Daytime_Sleepiness`, `Physical_Activity_Level`, and `Stress_Level` are discrete variables.

-   `Gender` is a binary variable.

### Univariate Distrubutions

```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(patchwork)

# ── Shared theme ──────────────────────────────────────────────────────────────
dist_theme <- theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 7)
  )

# ── Helper: per-variable binwidth + unit labels ───────────────────────────────
plot_faceted_binwidth_units <- function(data, var_binwidths, var_units, var_titles,
                                        title, subtitle, fill_color, ncol = 3) {
  vars  <- names(var_binwidths)
  plots <- map(vars, function(v) {
    bw         <- var_binwidths[[v]]
    panel_title <- var_titles[[v]]
    xlab       <- var_units[[v]]
    ggplot(data, aes(x = .data[[v]])) +
      geom_histogram(aes(y = after_stat(density)),
                     binwidth = bw, fill = fill_color,
                     color = "white", alpha = 0.8) +
      geom_density(color = "#C44E52", linewidth = 0.8) +
      labs(title = panel_title, x = xlab, y = "Density") +
      dist_theme +
      theme(
        plot.title   = element_text(face = "bold", size = 10),
        axis.title.x = element_text(size = 8, color = "grey40")
      )
  })

  wrap_plots(plots, ncol = ncol) +
    plot_annotation(
      title    = title,
      subtitle = subtitle,
      theme    = theme(
        plot.title    = element_text(face = "bold", size = 14),
        plot.subtitle = element_text(color = "grey50", size = 11)
      )
    )
}

# ── 1. Sleep Variables ────────────────────────────────────────────────────────
sleep_binwidths <- c(
  Sleep_Hours         = 0.5,
  Sleep_Quality_Score = 2,
  Daytime_Sleepiness  = 2
)
sleep_titles <- c(
  Sleep_Hours         = "Sleep Hours",
  Sleep_Quality_Score = "Sleep Quality Score",
  Daytime_Sleepiness  = "Daytime Sleepiness"
)
sleep_units <- c(
  Sleep_Hours         = "Duration (hours)",
  Sleep_Quality_Score = "Quality Score (0–20)",
  Daytime_Sleepiness  = "Sleepiness Score (0–25)"
)

p_sleep <- plot_faceted_binwidth_units(
  data          = df,
  var_binwidths = sleep_binwidths,
  var_units     = sleep_units,
  var_titles    = sleep_titles,
  title         = "Distributions of Sleep Variables",
  subtitle      = "Histogram with density overlay  |  n = 60",
  fill_color    = "#4C72B0",
  ncol          = 3
)
p_sleep

# ── 2. Cognitive Performance Variables ───────────────────────────────────────
cognitive_binwidths <- c(
  Stroop_Task_Reaction_Time = 0.25,
  N_Back_Accuracy           = 5,
  Emotion_Regulation_Score  = 5,
  PVT_Reaction_Time         = 25
)
cognitive_titles <- c(
  Stroop_Task_Reaction_Time = "Stroop Task Reaction Time",
  N_Back_Accuracy           = "N-Back Accuracy",
  Emotion_Regulation_Score  = "Emotion Regulation Score",
  PVT_Reaction_Time         = "PVT Reaction Time"
)
cognitive_units <- c(
  Stroop_Task_Reaction_Time = "Reaction Time (seconds)",
  N_Back_Accuracy           = "Accuracy (% correct)",
  Emotion_Regulation_Score  = "Score (0–70)",
  PVT_Reaction_Time         = "Reaction Time (milliseconds)"
)

p_cognitive <- plot_faceted_binwidth_units(
  data          = df,
  var_binwidths = cognitive_binwidths,
  var_units     = cognitive_units,
  var_titles    = cognitive_titles,
  title         = "Distributions of Cognitive Performance Variables",
  subtitle      = "Histogram with density overlay  |  n = 60",
  fill_color    = "#55A868",
  ncol          = 2
)
p_cognitive

# ── 3. Lifestyle Covariates (Continuous) ─────────────────────────────────────
lifestyle_binwidths <- c(
  Age          = 2,
  BMI          = 1.5,
  Stress_Level = 4
)
lifestyle_titles <- c(
  Age          = "Age",
  BMI          = "BMI",
  Stress_Level = "Stress Level"
)
lifestyle_units <- c(
  Age          = "Age (years)",
  BMI          = "BMI (kg/m²)",
  Stress_Level = "Stress Score (0–40)"
)

p_lifestyle_cont <- plot_faceted_binwidth_units(
  data          = df,
  var_binwidths = lifestyle_binwidths,
  var_units     = lifestyle_units,
  var_titles    = lifestyle_titles,
  title         = "Lifestyle Covariates (Continuous)",
  subtitle      = "Histogram with density overlay  |  n = 60",
  fill_color    = "#DD8452",
  ncol          = 3
)
p_lifestyle_cont

# ── 4. Discrete / Categorical Covariates ─────────────────────────────────────
lifestyle_discrete_long <- bind_rows(
  df |> count(value = factor(Caffeine_Intake))         |> mutate(variable = "Caffeine Intake (cups/day)"),
  df |> count(value = factor(Physical_Activity_Level)) |> mutate(variable = "Physical Activity Level (1–10)"),
  df |> count(value = factor(Gender))                  |> mutate(variable = "Gender")
)

p_lifestyle_disc <- ggplot(lifestyle_discrete_long,
                           aes(x = value, y = n, fill = variable)) +
  geom_col(show.legend = FALSE, width = 0.6, alpha = 0.85) +
  geom_text(aes(label = n), vjust = -0.4, size = 3) +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title    = "Lifestyle Covariates (Discrete / Categorical)",
    subtitle = "Count per class  |  n = 60  |  Caffeine: 0–5 cups/day, Physical Activity: 1–10 scale",
    x        = NULL,
    y        = "Count"
  ) +
  dist_theme +
  theme(panel.grid.major.x = element_blank())

# ── Remove individual titles from sub-plots before combining ─────────────────
p_lifestyle_cont_notitle <- p_lifestyle_cont &
  theme(plot.title = element_blank(), plot.subtitle = element_blank())

p_lifestyle_disc_notitle <- p_lifestyle_disc +
  labs(title = NULL, subtitle = NULL)

# ── Combined figure ───────────────────────────────────────────────────────────
(p_lifestyle_cont_notitle / p_lifestyle_disc_notitle) +
  plot_annotation(
    title    = "Lifestyle Covariates  |  n = 60",
    subtitle = "Top: continuous variables (histograms with density overlay)\nBottom: discrete and categorical variables (counts per class)",
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(color = "grey50", size = 11)
    )
  ) +
  plot_layout(heights = c(1.2, 1))
```

### Correlation Matrix

```{r corrplot}
library(tidyverse)
library(ggcorrplot)

numeric_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time",
  "Age", "BMI", "Caffeine_Intake", "Physical_Activity_Level", "Stress_Level"
)

cor_long <- df |>
  select(all_of(numeric_vars)) |>
  rename_with(~ str_replace_all(.x, "_", "\n")) |>
  cor(use = "pairwise.complete.obs") |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  # keep lower triangle only
  mutate(var1 = factor(var1, levels = unique(var1)),
         var2 = factor(var2, levels = unique(var2))) |>
  filter(as.integer(var1) > as.integer(var2))

p_corr <- ggplot(cor_long, aes(x = var2, y = var1, size = abs(r), fill = r)) +
  geom_point(shape = 21, color = "white") +
  geom_text(aes(label = round(r, 2)), size = 3, color = "grey20") +
  scale_fill_gradient2(low = "#C44E52", mid = "white", high = "#4C72B0",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  scale_size_continuous(range = c(2, 14), guide = "none") +
  labs(
    title    = "Correlation Matrix",
    subtitle = "Circle size = |r|",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "grey50", size = 11),
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7, lineheight = 0.85),
    axis.text.y     = element_text(size = 7, lineheight = 0.85),
    panel.grid      = element_blank(),
    legend.position = "right",
    plot.margin     = margin(10, 10, 10, 30)
  )

p_corr
```

### Scatterplots

```{r scatter_sleep_vs_cognition}
library(tidyverse)
library(ggplot2)
library(dplyr)

plot_sleep_vs_cognition <- function(data, sleep_var, sleep_label, sleep_unit) {

  cognitive_vars <- c(
    "Stroop_Task_Reaction_Time",
    "N_Back_Accuracy",
    "Emotion_Regulation_Score",
    "PVT_Reaction_Time"
  )

  # Descriptive panel titles with units
  outcome_labels <- c(
    "Stroop_Task_Reaction_Time" = "Stroop Task Reaction Time (seconds)",
    "N_Back_Accuracy"           = "N-Back Accuracy (% correct)",
    "Emotion_Regulation_Score"  = "Emotion Regulation Score (0–70)",
    "PVT_Reaction_Time"         = "PVT Reaction Time (milliseconds)"
  )

  data |>
    select(all_of(c(sleep_var, cognitive_vars))) |>
    pivot_longer(-all_of(sleep_var), names_to = "outcome", values_to = "value") |>
    mutate(outcome = recode(outcome, !!!outcome_labels)) |>
    ggplot(aes(x = .data[[sleep_var]], y = value)) +
    geom_point(alpha = 0.4, size = 1.8, color = "#4C72B0") +
    geom_smooth(method = "lm", se = TRUE, color = "#C44E52", linewidth = 1) +
    facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
    labs(
      title    = paste("Sleep vs. Cognitive Outcomes:", sleep_label),
      subtitle = "Linear fit with 95% confidence interval",
      x        = sleep_unit,
      y        = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey50", size = 11),
      strip.text       = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank()
    )
}

# ── Generate both plots ───────────────────────────────────────────────────────
p_hours   <- plot_sleep_vs_cognition(
  data        = df,
  sleep_var   = "Sleep_Hours",
  sleep_label = "Sleep Hours",
  sleep_unit  = "Sleep Duration (hours)"
)

p_quality <- plot_sleep_vs_cognition(
  data        = df,
  sleep_var   = "Sleep_Quality_Score",
  sleep_label = "Sleep Quality Score",
  sleep_unit  = "Sleep Quality Score (0–20)"
)

print(p_hours)
print(p_quality)
```

### K-Means

```{r K-Means: behavioral}
# ─────────────────────────────────────────────
# K-Means Clustering: Behavioral Features
# ─────────────────────────────────────────────

library(tidyverse)
library(cluster)
library(factoextra)
library(patchwork)

# ── 1. Prep behavioral features ──────────────
behav_vars <- c("BMI", "Caffeine_Intake", "Physical_Activity_Level", "Stress_Level")

df_behav <- df %>%
  select(all_of(behav_vars)) %>%
  drop_na() %>%
  scale() %>%
  as.data.frame()

# ── 2. Elbow plot (WSS) ───────────────────────
set.seed(42)
wss <- map_dbl(1:10, function(k) {
  kmeans(df_behav, centers = k, nstart = 25, iter.max = 100)$tot.withinss
})

elbow_df <- tibble(k = 1:10, wss = wss)

p_elbow <- ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line(color = "#4C72B0", linewidth = 1) +
  geom_point(size = 3, color = "#4C72B0") +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Elbow Method: Optimal k",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster SS"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

print(p_elbow)

# ── 3. Fit final model (set k after inspecting elbow) ──
k_final <- 3  # ← adjust after viewing elbow plot

km_fit <- kmeans(df_behav, centers = k_final, nstart = 25, iter.max = 100)

# Attach cluster labels back to original (unscaled) rows
df_clustered <- df %>%
  drop_na(all_of(behav_vars)) %>%
  mutate(cluster = factor(km_fit$cluster))

# ── 4. PCA scatter ────────────────────────────
pca_out  <- prcomp(df_behav)
pca_df   <- as.data.frame(pca_out$x[, 1:2]) %>%
  mutate(cluster = factor(km_fit$cluster))

var_exp  <- round(summary(pca_out)$importance[2, 1:2] * 100, 1)

p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, color = cluster, fill = cluster)) +
  geom_point(size = 3, alpha = 0.8, shape = 21, color = "white", stroke = 0.4,
             aes(fill = cluster)) +
  stat_ellipse(geom = "polygon", alpha = 0.12, linewidth = 0.6) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "K-Means Clusters (PCA Space)",
    x = paste0("PC1 (", var_exp[1], "% variance)"),
    y = paste0("PC2 (", var_exp[2], "% variance)"),
    fill = "Cluster", color = "Cluster"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

# ── 5. Faceted boxplots by cluster ───────────
df_long <- df_clustered %>%
  select(all_of(behav_vars), cluster) %>%
  pivot_longer(cols = all_of(behav_vars),
               names_to = "feature",
               values_to = "value") %>%
  mutate(feature = recode(feature,
    "BMI"                   = "BMI",
    "Caffeine_Intake"       = "Caffeine Intake",
    "Physical_Activity_Level" = "Physical Activity",
    "Stress_Level"          = "Stress Level"
  ))

p_box <- ggplot(df_long, aes(x = cluster, y = value, fill = cluster)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  facet_wrap(~ feature, scales = "free_y", ncol = 2) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Behavioral Features by Cluster",
    x = "Cluster", y = "Value", fill = "Cluster"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title   = element_text(face = "bold"),
    legend.position = "none"
  )

# ── 6. Combined layout ────────────────────────
(p_pca | p_box) +
  plot_annotation(
    title   = "K-Means Clustering: Behavioral Health Profiles",
    subtitle = paste0("k = ", k_final, " | Features: BMI, Caffeine, Physical Activity, Stress"),
    theme   = theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(size = 11, color = "grey40")
    )
  )

# ── 7. Cluster summary table ──────────────────
df_clustered %>%
  group_by(cluster) %>%
  summarise(
    n                = n(),
    BMI_mean         = round(mean(BMI, na.rm = TRUE), 2),
    Caffeine_mean    = round(mean(Caffeine_Intake, na.rm = TRUE), 2),
    PhysActivity_mean = round(mean(Physical_Activity_Level, na.rm = TRUE), 2),
    Stress_mean      = round(mean(Stress_Level, na.rm = TRUE), 2)
  ) %>%
  print()
```

```{r}
# ─────────────────────────────────────────────
# K-Means Clustering: Sleep + Cognitive Features
# ─────────────────────────────────────────────

library(tidyverse)
library(patchwork)

# ── 1. Prep sleep + cognitive features ───────
sleep_cog_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "PVT_Reaction_Time", "Emotion_Regulation_Score"
)

df_sc <- df %>%
  select(all_of(sleep_cog_vars)) %>%
  drop_na() %>%
  scale() %>%
  as.data.frame()

# ── 2. Elbow plot (WSS) ───────────────────────
set.seed(42)
wss <- map_dbl(1:10, function(k) {
  kmeans(df_sc, centers = k, nstart = 25, iter.max = 100)$tot.withinss
})

elbow_df <- tibble(k = 1:10, wss = wss)

p_elbow <- ggplot(elbow_df, aes(x = k, y = wss)) +
  geom_line(color = "#DD8452", linewidth = 1) +
  geom_point(size = 3, color = "#DD8452") +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Elbow Method: Optimal k",
    x = "Number of Clusters (k)",
    y = "Total Within-Cluster SS"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

print(p_elbow)

# ── 3. Fit final model (set k after inspecting elbow) ──
k_final <- 3  # ← adjust after viewing elbow plot

km_fit <- kmeans(df_sc, centers = k_final, nstart = 25, iter.max = 100)

# Attach cluster labels back to original (unscaled) rows
df_clustered_sc <- df %>%
  drop_na(all_of(sleep_cog_vars)) %>%
  mutate(cluster = factor(km_fit$cluster))

# ── 4. PCA scatter ────────────────────────────
pca_out <- prcomp(df_sc)
pca_df  <- as.data.frame(pca_out$x[, 1:2]) %>%
  mutate(cluster = factor(km_fit$cluster))

var_exp <- round(summary(pca_out)$importance[2, 1:2] * 100, 1)

# PCA loadings for reference
loadings_df <- as.data.frame(pca_out$rotation[, 1:2]) %>%
  rownames_to_column("variable") %>%
  mutate(variable = recode(variable,
    "Sleep_Hours"               = "Sleep Hours",
    "Sleep_Quality_Score"       = "Sleep Quality",
    "Daytime_Sleepiness"        = "Daytime Sleepiness",
    "Stroop_Task_Reaction_Time" = "Stroop RT",
    "N_Back_Accuracy"           = "N-Back Accuracy",
    "PVT_Reaction_Time"         = "PVT RT",
    "Emotion_Regulation_Score"  = "Emotion Regulation"
  ))

print(loadings_df)  # inspect which variables drive each PC

p_pca <- ggplot(pca_df, aes(x = PC1, y = PC2, fill = cluster)) +
  geom_point(size = 3, alpha = 0.9, shape = 21, color = "white", stroke = 0.4) +
  stat_ellipse(geom = "polygon", alpha = 0.12, aes(color = cluster), linewidth = 0.6) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  labs(
    title = "K-Means Clusters (PCA Space)",
    x = paste0("PC1 (", var_exp[1], "% variance)"),
    y = paste0("PC2 (", var_exp[2], "% variance)"),
    fill = "Cluster", color = "Cluster"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

# ── 5. Faceted boxplots by cluster ───────────
df_long_sc <- df_clustered_sc %>%
  select(all_of(sleep_cog_vars), cluster) %>%
  pivot_longer(cols = all_of(sleep_cog_vars),
               names_to = "feature",
               values_to = "value") %>%
  mutate(feature = recode(feature,
    "Sleep_Hours"               = "Sleep Hours",
    "Sleep_Quality_Score"       = "Sleep Quality",
    "Daytime_Sleepiness"        = "Daytime Sleepiness",
    "Stroop_Task_Reaction_Time" = "Stroop RT",
    "N_Back_Accuracy"           = "N-Back Accuracy",
    "PVT_Reaction_Time"         = "PVT RT",
    "Emotion_Regulation_Score"  = "Emotion Regulation"
  ))

p_box <- ggplot(df_long_sc, aes(x = cluster, y = value, fill = cluster)) +
  geom_boxplot(alpha = 0.7, outlier.shape = 21, outlier.size = 2) +
  geom_jitter(width = 0.15, alpha = 0.4, size = 1.5) +
  facet_wrap(~ feature, scales = "free_y", ncol = 3) +
  scale_fill_brewer(palette = "Set2") +
  labs(
    title = "Sleep & Cognitive Features by Cluster",
    x = "Cluster", y = "Value", fill = "Cluster"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "none"
  )

# ── 6. Combined layout ────────────────────────
(p_pca | p_box) +
  plot_annotation(
    title    = "K-Means Clustering: Sleep & Cognitive Profiles",
    subtitle = paste0("k = ", k_final, " | Features: Sleep Hours, Sleep Quality, Daytime Sleepiness, Stroop RT, N-Back Accuracy, PVT RT, Emotion Regulation"),
    theme    = theme(
      plot.title    = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(size = 10, color = "grey40")
    )
  )

# ── 7. Cluster summary table ──────────────────
df_clustered_sc %>%
  group_by(cluster) %>%
  summarise(
    n                      = n(),
    Sleep_Hours_mean       = round(mean(Sleep_Hours, na.rm = TRUE), 2),
    Sleep_Quality_mean     = round(mean(Sleep_Quality_Score, na.rm = TRUE), 2),
    Daytime_Sleep_mean     = round(mean(Daytime_Sleepiness, na.rm = TRUE), 2),
    Stroop_RT_mean         = round(mean(Stroop_Task_Reaction_Time, na.rm = TRUE), 2),
    NBack_Acc_mean         = round(mean(N_Back_Accuracy, na.rm = TRUE), 2),
    PVT_RT_mean            = round(mean(PVT_Reaction_Time, na.rm = TRUE), 2),
    Emotion_Reg_mean       = round(mean(Emotion_Regulation_Score, na.rm = TRUE), 2)
  ) %>%
  print()
```

## Models

### Stroop Reaction Time

```{r}
library(tidyverse)
library(glmnet)
library(broom)
library(patchwork)

# ── Data prep ─────────────────────────────────
df_model <- df %>%
  select(
    Stroop_Task_Reaction_Time,
    Sleep_Hours, Sleep_Quality_Score, Daytime_Sleepiness,
    BMI, Caffeine_Intake, Physical_Activity_Level,
    Stress_Level, Age, Gender
  ) %>%
  drop_na() %>%
  mutate(Gender = factor(Gender))

# ── Model 1: Unadjusted OLS (sleep only) ─────
m1 <- lm(Stroop_Task_Reaction_Time ~
           Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness,
         data = df_model)

cat("\n========== Model 1: Unadjusted (Sleep Only) ==========\n")
summary(m1)

# ── Model 2: Adjusted OLS (sleep + covariates) 
m2 <- lm(Stroop_Task_Reaction_Time ~
           Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness +
           BMI + Caffeine_Intake + Physical_Activity_Level +
           Stress_Level + Age + Gender,
         data = df_model)

cat("\n========== Model 2: Adjusted (Sleep + Covariates) ==========\n")
summary(m2)

# ── Model 3: LASSO ────────────────────────────
X <- model.matrix(Stroop_Task_Reaction_Time ~
                    Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness +
                    BMI + Caffeine_Intake + Physical_Activity_Level +
                    Stress_Level + Age + Gender,
                  data = df_model)[, -1]  # drop intercept

y <- df_model$Stroop_Task_Reaction_Time

set.seed(42)
lasso_cv <- cv.glmnet(X, y, alpha = 1, nfolds = 10)

cat("\n========== Model 3: LASSO ==========\n")
cat("Optimal lambda (lambda.min):", lasso_cv$lambda.min, "\n")
cat("Lambda 1SE:", lasso_cv$lambda.1se, "\n\n")

lasso_coefs <- coef(lasso_cv, s = "lambda.1se")
print(lasso_coefs)

# ── Residual diagnostics for OLS models ──────
par(mfrow = c(2, 2))
plot(m1, main = "Model 1 Diagnostics")

par(mfrow = c(2, 2))
plot(m2, main = "Model 2 Diagnostics")

# ── Coefficient plot: Model 1 vs Model 2 ─────
tidy_m1 <- tidy(m1, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Unadjusted")

tidy_m2 <- tidy(m2, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Adjusted")

coef_df <- bind_rows(tidy_m1, tidy_m2) %>%
  mutate(term = recode(term,
    "Sleep_Hours"              = "Sleep Hours",
    "Sleep_Quality_Score"      = "Sleep Quality Score",
    "Daytime_Sleepiness"       = "Daytime Sleepiness",
    "BMI"                      = "BMI",
    "Caffeine_Intake"          = "Caffeine Intake",
    "Physical_Activity_Level"  = "Physical Activity Level",
    "Stress_Level"             = "Stress Level",
    "Age"                      = "Age",
    "GenderMale"               = "Gender (Male)"
  ))

ggplot(coef_df, aes(x = estimate, y = reorder(term, estimate), color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, position = position_dodge(0.5)) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "OLS Coefficients: Stroop Task Reaction Time",
    subtitle = "Unadjusted vs. adjusted model with 95% confidence intervals",
    x        = "Coefficient Estimate (seconds)",
    y        = NULL,
    color    = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(color = "grey50"),
    legend.position = "bottom"
  )

# ── LASSO path plot ───────────────────────────
plot(lasso_cv,
     main = "LASSO Cross-Validation: Stroop Task Reaction Time")
```

### PVT Reaction Time

```{r}
# ══════════════════════════════════════════════
# PVT Reaction Time
# ══════════════════════════════════════════════

df_model_pvt <- df %>%
  select(
    PVT_Reaction_Time,
    Sleep_Hours, Sleep_Quality_Score, Daytime_Sleepiness,
    BMI, Caffeine_Intake, Physical_Activity_Level,
    Stress_Level, Age, Gender
  ) %>%
  drop_na() %>%
  mutate(Gender = factor(Gender))

# ── Model 1: Unadjusted OLS (sleep only) ─────
m1_pvt <- lm(PVT_Reaction_Time ~
               Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness,
             data = df_model_pvt)

cat("\n========== PVT RT | Model 1: Unadjusted (Sleep Only) ==========\n")
summary(m1_pvt)

# ── Model 2: Adjusted OLS ─────────────────────
m2_pvt <- lm(PVT_Reaction_Time ~
               Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness +
               BMI + Caffeine_Intake + Physical_Activity_Level +
               Stress_Level + Age + Gender,
             data = df_model_pvt)

cat("\n========== PVT RT | Model 2: Adjusted (Sleep + Covariates) ==========\n")
summary(m2_pvt)

# ── Model 3: LASSO ────────────────────────────
X_pvt <- model.matrix(PVT_Reaction_Time ~
                        Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness +
                        BMI + Caffeine_Intake + Physical_Activity_Level +
                        Stress_Level + Age + Gender,
                      data = df_model_pvt)[, -1]

y_pvt <- df_model_pvt$PVT_Reaction_Time

set.seed(42)
lasso_cv_pvt <- cv.glmnet(X_pvt, y_pvt, alpha = 1, nfolds = 10)

cat("\n========== PVT RT | Model 3: LASSO ==========\n")
cat("Optimal lambda (lambda.min):", lasso_cv_pvt$lambda.min, "\n")
cat("Lambda 1SE:", lasso_cv_pvt$lambda.1se, "\n\n")
print(coef(lasso_cv_pvt, s = "lambda.1se"))

# ── Residual diagnostics ──────────────────────
par(mfrow = c(2, 2))
plot(m1_pvt, main = "PVT RT | Model 1 Diagnostics")

par(mfrow = c(2, 2))
plot(m2_pvt, main = "PVT RT | Model 2 Diagnostics")

# ── Coefficient plot ──────────────────────────
tidy_m1_pvt <- tidy(m1_pvt, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Unadjusted")

tidy_m2_pvt <- tidy(m2_pvt, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Adjusted")

coef_df_pvt <- bind_rows(tidy_m1_pvt, tidy_m2_pvt) %>%
  mutate(term = recode(term,
    "Sleep_Hours"             = "Sleep Hours",
    "Sleep_Quality_Score"     = "Sleep Quality Score",
    "Daytime_Sleepiness"      = "Daytime Sleepiness",
    "BMI"                     = "BMI",
    "Caffeine_Intake"         = "Caffeine Intake",
    "Physical_Activity_Level" = "Physical Activity Level",
    "Stress_Level"            = "Stress Level",
    "Age"                     = "Age",
    "GenderMale"              = "Gender (Male)"
  ))

ggplot(coef_df_pvt, aes(x = estimate, y = reorder(term, estimate), color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, position = position_dodge(0.5)) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "OLS Coefficients: PVT Reaction Time",
    subtitle = "Unadjusted vs. adjusted model with 95% confidence intervals",
    x        = "Coefficient Estimate (milliseconds)",
    y        = NULL,
    color    = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey50"),
    legend.position = "bottom"
  )

# ── LASSO CV plot ─────────────────────────────
plot(lasso_cv_pvt,
     main = "LASSO Cross-Validation: PVT Reaction Time")
```

### N-Back Accuracy

```{r}
# ══════════════════════════════════════════════
# N-Back Accuracy
# ══════════════════════════════════════════════

df_model_nback <- df %>%
  select(
    N_Back_Accuracy,
    Sleep_Hours, Sleep_Quality_Score, Daytime_Sleepiness,
    BMI, Caffeine_Intake, Physical_Activity_Level,
    Stress_Level, Age, Gender
  ) %>%
  drop_na() %>%
  mutate(Gender = factor(Gender))

# ── Model 1: Unadjusted OLS (sleep only) ─────
m1_nback <- lm(N_Back_Accuracy ~
                 Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness,
               data = df_model_nback)

cat("\n========== N-Back Accuracy | Model 1: Unadjusted (Sleep Only) ==========\n")
summary(m1_nback)

# ── Model 2: Adjusted OLS ─────────────────────
m2_nback <- lm(N_Back_Accuracy ~
                 Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness +
                 BMI + Caffeine_Intake + Physical_Activity_Level +
                 Stress_Level + Age + Gender,
               data = df_model_nback)

cat("\n========== N-Back Accuracy | Model 2: Adjusted (Sleep + Covariates) ==========\n")
summary(m2_nback)

# ── Model 3: LASSO ────────────────────────────
X_nback <- model.matrix(N_Back_Accuracy ~
                          Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness +
                          BMI + Caffeine_Intake + Physical_Activity_Level +
                          Stress_Level + Age + Gender,
                        data = df_model_nback)[, -1]

y_nback <- df_model_nback$N_Back_Accuracy

set.seed(42)
lasso_cv_nback <- cv.glmnet(X_nback, y_nback, alpha = 1, nfolds = 10)

cat("\n========== N-Back Accuracy | Model 3: LASSO ==========\n")
cat("Optimal lambda (lambda.min):", lasso_cv_nback$lambda.min, "\n")
cat("Lambda 1SE:", lasso_cv_nback$lambda.1se, "\n\n")
print(coef(lasso_cv_nback, s = "lambda.1se"))

# ── Residual diagnostics ──────────────────────
par(mfrow = c(2, 2))
plot(m1_nback, main = "N-Back Accuracy | Model 1 Diagnostics")

par(mfrow = c(2, 2))
plot(m2_nback, main = "N-Back Accuracy | Model 2 Diagnostics")

# ── Coefficient plot ──────────────────────────
tidy_m1_nback <- tidy(m1_nback, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Unadjusted")

tidy_m2_nback <- tidy(m2_nback, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Adjusted")

coef_df_nback <- bind_rows(tidy_m1_nback, tidy_m2_nback) %>%
  mutate(term = recode(term,
    "Sleep_Hours"             = "Sleep Hours",
    "Sleep_Quality_Score"     = "Sleep Quality Score",
    "Daytime_Sleepiness"      = "Daytime Sleepiness",
    "BMI"                     = "BMI",
    "Caffeine_Intake"         = "Caffeine Intake",
    "Physical_Activity_Level" = "Physical Activity Level",
    "Stress_Level"            = "Stress Level",
    "Age"                     = "Age",
    "GenderMale"              = "Gender (Male)"
  ))

ggplot(coef_df_nback, aes(x = estimate, y = reorder(term, estimate), color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, position = position_dodge(0.5)) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "OLS Coefficients: N-Back Accuracy",
    subtitle = "Unadjusted vs. adjusted model with 95% confidence intervals",
    x        = "Coefficient Estimate (% correct)",
    y        = NULL,
    color    = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey50"),
    legend.position = "bottom"
  )

# ── LASSO CV plot ─────────────────────────────
plot(lasso_cv_nback,
     main = "LASSO Cross-Validation: N-Back Accuracy")
```

### Emotion Regulation Score

```{r}
# ══════════════════════════════════════════════
# Emotion Regulation Score
# ══════════════════════════════════════════════

df_model_emo <- df %>%
  select(
    Emotion_Regulation_Score,
    Sleep_Hours, Sleep_Quality_Score, Daytime_Sleepiness,
    BMI, Caffeine_Intake, Physical_Activity_Level,
    Stress_Level, Age, Gender
  ) %>%
  drop_na() %>%
  mutate(Gender = factor(Gender))

# ── Model 1: Unadjusted OLS (sleep only) ─────
m1_emo <- lm(Emotion_Regulation_Score ~
               Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness,
             data = df_model_emo)

cat("\n========== Emotion Regulation | Model 1: Unadjusted (Sleep Only) ==========\n")
summary(m1_emo)

# ── Model 2: Adjusted OLS ─────────────────────
m2_emo <- lm(Emotion_Regulation_Score ~
               Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness +
               BMI + Caffeine_Intake + Physical_Activity_Level +
               Stress_Level + Age + Gender,
             data = df_model_emo)

cat("\n========== Emotion Regulation | Model 2: Adjusted (Sleep + Covariates) ==========\n")
summary(m2_emo)

# ── Model 3: LASSO ────────────────────────────
X_emo <- model.matrix(Emotion_Regulation_Score ~
                        Sleep_Hours + Sleep_Quality_Score + Daytime_Sleepiness +
                        BMI + Caffeine_Intake + Physical_Activity_Level +
                        Stress_Level + Age + Gender,
                      data = df_model_emo)[, -1]

y_emo <- df_model_emo$Emotion_Regulation_Score

set.seed(42)
lasso_cv_emo <- cv.glmnet(X_emo, y_emo, alpha = 1, nfolds = 10)

cat("\n========== Emotion Regulation | Model 3: LASSO ==========\n")
cat("Optimal lambda (lambda.min):", lasso_cv_emo$lambda.min, "\n")
cat("Lambda 1SE:", lasso_cv_emo$lambda.1se, "\n\n")
print(coef(lasso_cv_emo, s = "lambda.1se"))

# ── Residual diagnostics ──────────────────────
par(mfrow = c(2, 2))
plot(m1_emo, main = "Emotion Regulation | Model 1 Diagnostics")

par(mfrow = c(2, 2))
plot(m2_emo, main = "Emotion Regulation | Model 2 Diagnostics")

# ── Coefficient plot ──────────────────────────
tidy_m1_emo <- tidy(m1_emo, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Unadjusted")

tidy_m2_emo <- tidy(m2_emo, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(model = "Adjusted")

coef_df_emo <- bind_rows(tidy_m1_emo, tidy_m2_emo) %>%
  mutate(term = recode(term,
    "Sleep_Hours"             = "Sleep Hours",
    "Sleep_Quality_Score"     = "Sleep Quality Score",
    "Daytime_Sleepiness"      = "Daytime Sleepiness",
    "BMI"                     = "BMI",
    "Caffeine_Intake"         = "Caffeine Intake",
    "Physical_Activity_Level" = "Physical Activity Level",
    "Stress_Level"            = "Stress Level",
    "Age"                     = "Age",
    "GenderMale"              = "Gender (Male)"
  ))

ggplot(coef_df_emo, aes(x = estimate, y = reorder(term, estimate), color = model)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey50") +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high),
                 height = 0.25, position = position_dodge(0.5)) +
  geom_point(size = 3, position = position_dodge(0.5)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title    = "OLS Coefficients: Emotion Regulation Score",
    subtitle = "Unadjusted vs. adjusted model with 95% confidence intervals",
    x        = "Coefficient Estimate (score units, 0–70)",
    y        = NULL,
    color    = "Model"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    plot.subtitle   = element_text(color = "grey50"),
    legend.position = "bottom"
  )

# ── LASSO CV plot ─────────────────────────────
plot(lasso_cv_emo,
     main = "LASSO Cross-Validation: Emotion Regulation Score")
```
