---
title: "Sleep Deprivation and Cognitive Performance"
format: html
editor: visual
---

## Sleep Deprivation and Cognitive Performance

## Goals

-   Understand the effects of sleep deprivation (hours of sleep, sleep quality, daytime sleep) and behavioral determinants of health (BMI, caffeine intake, physical activity level) on cognitive performance (attention, working memory, reaction speed, emotion regulation).
-   Model each cognitive performance metric using sleep and behavioral health features.

## Read Data

```{r read_data}
library(readr)

df <- read_csv("sleep_deprivation_dataset_detailed.csv")
head(df)
```

## Exploratory Data Analysis (EDA)

### Data Structure

```{r data_structure}
str(df)
```

### Summary Statistics

```{r summary}
summary(df)
```

```{r unique_gender}
cat("Unique values of df$Gender:", unique(df$Gender), "\n")
cat("Unique values of df$Caffeine_Intake", sort(unique(df$Caffeine_Intake)), "\n")
```

Note:

-   `Caffeine_Intake`, `Sleep_Quality_Score`, `Daytime_Sleepiness`, `Physical_Activity_Level`, and `Stress_Level` are discrete variables.

-   `Gender` is a binary variable.

### Univariate Distrubutions

```{r continuous_dists}
library(ggplot2)
library(tidyverse)
library(dplyr)

continuous_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time",
  "Age", "BMI", "Stress_Level"
)

df_long_cont <- df |>
  select(all_of(continuous_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(variable = str_replace_all(variable, "_", " "))

p_continuous <- ggplot(df_long_cont, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "#4C72B0", color = "white", alpha = 0.8) +
  geom_density(color = "#C44E52", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(
    title    = "Univariate Distributions — Continuous Variables",
    subtitle = "Histogram with density overlay",
    x = NULL, y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 7)
  )

p_continuous
```

```{r discrete_dists}
library(ggplot2)
library(tidyverse)
library(dplyr)

discrete_long <- bind_rows(
  df |> count(value = factor(Caffeine_Intake))       |> mutate(variable = "Caffeine Intake (1–5)"),
  df |> count(value = factor(Physical_Activity_Level)) |> mutate(variable = "Physical Activity Level (1–10)"),
  df |> count(value = factor(Gender))                |> mutate(variable = "Gender")
)

p_discrete <- ggplot(discrete_long, aes(x = value, y = n, fill = variable)) +
  geom_col(show.legend = FALSE, width = 0.6, alpha = 0.85) +
  geom_text(aes(label = n), vjust = -0.4, size = 3) +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title    = "Univariate Distributions — Discrete & Categorical Variables",
    subtitle = "Count per class",
    x = NULL, y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title        = element_text(face = "bold", size = 14),
    plot.subtitle     = element_text(color = "grey50", size = 11),
    strip.text        = element_text(face = "bold", size = 9),
    panel.grid.major.x = element_blank(),
    panel.grid.minor  = element_blank()
  )

p_discrete
```

### Correlation Matrix

```{r corrplot}
library(tidyverse)
library(ggcorrplot)

numeric_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time",
  "Age", "BMI", "Caffeine_Intake", "Physical_Activity_Level", "Stress_Level"
)

cor_long <- df |>
  select(all_of(numeric_vars)) |>
  rename_with(~ str_replace_all(.x, "_", "\n")) |>
  cor(use = "pairwise.complete.obs") |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  # keep lower triangle only
  mutate(var1 = factor(var1, levels = unique(var1)),
         var2 = factor(var2, levels = unique(var2))) |>
  filter(as.integer(var1) > as.integer(var2))

p_corr <- ggplot(cor_long, aes(x = var2, y = var1, size = abs(r), fill = r)) +
  geom_point(shape = 21, color = "white") +
  geom_text(aes(label = round(r, 2)), size = 3, color = "grey20") +
  scale_fill_gradient2(low = "#C44E52", mid = "white", high = "#4C72B0",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  scale_size_continuous(range = c(2, 14), guide = "none") +
  labs(
    title    = "Correlation Matrix",
    subtitle = "Circle size = |r|",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "grey50", size = 11),
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7, lineheight = 0.85),
    axis.text.y     = element_text(size = 7, lineheight = 0.85),
    panel.grid      = element_blank(),
    legend.position = "right",
    plot.margin     = margin(10, 10, 10, 30)
  )

p_corr
```

### Scatterplots

```{r scatter_sleep_vs_cognition}
library(tidyverse)
library(ggplot2)
library(dplyr)

plot_sleep_vs_cognition <- function(data, sleep_var, sleep_label) {
  
  cognitive_vars <- c(
    "Stroop_Task_Reaction_Time",
    "N_Back_Accuracy",
    "Emotion_Regulation_Score",
    "PVT_Reaction_Time"
  )
  
  data |>
    select(all_of(c(sleep_var, cognitive_vars))) |>
    pivot_longer(-all_of(sleep_var), names_to = "outcome", values_to = "value") |>
    mutate(outcome = str_replace_all(outcome, "_", " ")) |>
    ggplot(aes(x = .data[[sleep_var]], y = value)) +
    geom_point(alpha = 0.4, size = 1.8, color = "#4C72B0") +
    geom_smooth(method = "lm", se = TRUE, color = "#C44E52", linewidth = 1) +
    facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
    labs(
      title    = paste("Sleep vs. Cognitive Outcomes —", sleep_label),
      subtitle = "Linear fit with 95% confidence interval",
      x        = sleep_label,
      y        = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title     = element_text(face = "bold", size = 15),
      plot.subtitle  = element_text(color = "grey50", size = 11),
      strip.text     = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank()
    )
}

# ── Generate both plots ───────────────────────────────────────────────────────
p_hours   <- plot_sleep_vs_cognition(df, "Sleep_Hours",         "Sleep Hours")
p_quality <- plot_sleep_vs_cognition(df, "Sleep_Quality_Score", "Sleep Quality Score")

print(p_hours)
print(p_quality)
```

### Binned Scatterplots

```{r binned_scatterplots}
library(tidyverse)
library(ggplot2)
library(dplyr)

df <- df |>
  mutate(Age_Group = cut(Age,
                         breaks = c(-Inf, 25, 35, 45, Inf),
                         labels = c("≤25", "26–35", "36–45", "46+")))

plot_binned <- function(data, sleep_var, sleep_label, group_var, group_label) {
  
  cognitive_vars <- c(
    "Stroop_Task_Reaction_Time",
    "N_Back_Accuracy",
    "Emotion_Regulation_Score",
    "PVT_Reaction_Time"
  )
  
  data |>
    select(all_of(c(sleep_var, group_var, cognitive_vars))) |>
    pivot_longer(-all_of(c(sleep_var, group_var)),
                 names_to = "outcome", values_to = "value") |>
    mutate(outcome = str_replace_all(outcome, "_", " ")) |>
    ggplot(aes(x = .data[[sleep_var]], y = value,
               color = .data[[group_var]], fill = .data[[group_var]])) +
    geom_point(alpha = 0.3, size = 1.5) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 1, alpha = 0.15) +
    facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
    labs(
      title    = paste("Sleep vs. Cognitive Outcomes —", sleep_label),
      subtitle = paste("By", group_label, "| Linear fit with 95% CI"),
      x        = sleep_label,
      y        = NULL,
      color    = group_label,
      fill     = group_label
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey50", size = 11),
      strip.text       = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
}


p_hours_gender   <- plot_binned(df, "Sleep_Hours",         "Sleep Hours",         "Gender", "Gender")
p_quality_gender <- plot_binned(df, "Sleep_Quality_Score", "Sleep Quality Score", "Gender", "Gender")

p_hours_age   <- plot_binned(df, "Sleep_Hours",         "Sleep Hours",         "Age_Group", "Age Group")
p_quality_age <- plot_binned(df, "Sleep_Quality_Score", "Sleep Quality Score", "Age_Group", "Age Group")


gender_colors <- c("#4C72B0", "#C44E52", "#55A868")  # flexible for 2–3 gender categories
age_colors    <- c("#4C72B0", "#55A868", "#C44E52", "#8172B2")

p_hours_gender   <- p_hours_gender   + scale_color_manual(values = gender_colors) + scale_fill_manual(values = gender_colors)
p_quality_gender <- p_quality_gender + scale_color_manual(values = gender_colors) + scale_fill_manual(values = gender_colors)
p_hours_age      <- p_hours_age      + scale_color_manual(values = age_colors)    + scale_fill_manual(values = age_colors)
p_quality_age    <- p_quality_age    + scale_color_manual(values = age_colors)    + scale_fill_manual(values = age_colors)

print(p_hours_gender)
print(p_quality_gender)
print(p_hours_age)
print(p_quality_age)
```

## Clustering

```{r cluster_by_sleep_group}
library(tidyverse)
library(dplyr)

df <- df |>
  mutate(Sleep_Group = cut(Sleep_Hours,
                           breaks = c(-Inf, 5, 6, 7, 9, Inf),
                           labels = c("Minimal\n(<5h)", "Poor\n(5–6h)",
                                      "Decent\n(6–7h)", "Sufficient\n(7–9h)",
                                      "Excessive\n(>9h)")))

# ── Pivot cognitive outcomes to long format ───────────────────────────────────
cognitive_vars <- c(
  "Stroop_Task_Reaction_Time",
  "N_Back_Accuracy",
  "Emotion_Regulation_Score",
  "PVT_Reaction_Time"
)

df_long <- df |>
  select(Sleep_Hours, Sleep_Group, all_of(cognitive_vars)) |>
  pivot_longer(-c(Sleep_Hours, Sleep_Group),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome = str_replace_all(outcome, "_", " "))

# ── Color palette: intuitive red → green spectrum ────────────────────────────
sleep_colors <- c(
  "Minimal\n(<5h)"    = "#C44E52",
  "Poor\n(5–6h)"      = "#DD8452",
  "Decent\n(6–7h)"    = "#CCB974",
  "Sufficient\n(7–9h)"= "#55A868",
  "Excessive\n(>9h)"  = "#4C72B0"
)

# ── Plot ──────────────────────────────────────────────────────────────────────
p <- ggplot(df_long, aes(x = Sleep_Hours, y = value,
                         color = Sleep_Group, fill = Sleep_Group)) +
  geom_point(alpha = 0.4, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, alpha = 0.15) +
  scale_color_manual(values = sleep_colors, name = "Sleep Duration") +
  scale_fill_manual(values  = sleep_colors, name = "Sleep Duration") +
  facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
  labs(
    title    = "Sleep Hours vs. Cognitive Outcomes by Sleep Duration Group",
    subtitle = "Each group has its own linear fit | Points colored by sleep duration bin",
    x        = "Sleep Hours",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold"),
    legend.text      = element_text(size = 9)
  ) +
  guides(color = guide_legend(nrow = 1, override.aes = list(alpha = 1, size = 3)),
         fill  = guide_legend(nrow = 1))

p
```

```{r}
cognitive_vars <- c(
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score",  "PVT_Reaction_Time"
)

df_scaled <- df |>
  select(all_of(cognitive_vars)) |>
  scale() |>
  as.data.frame()

# ══════════════════════════════════════════════════════════════════════════════
# 2. ELBOW PLOT — choose optimal k
# ══════════════════════════════════════════════════════════════════════════════
set.seed(42)
wss <- map_dbl(1:8, ~ kmeans(df_scaled, centers = .x, nstart = 25)$tot.withinss)

p_elbow <- tibble(k = 1:8, wss = wss) |>
  ggplot(aes(x = k, y = wss)) +
  geom_line(linewidth = 1, color = "grey50") +
  geom_point(size = 4, color = "#4C72B0") +
  scale_x_continuous(breaks = 1:8) +
  labs(
    title    = "Elbow Plot — K-Means on Cognitive Outcomes",
    subtitle = "Look for the 'elbow' where WSS stops dropping sharply",
    x        = "Number of Clusters (k)",
    y        = "Total Within-Cluster SS"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey50", size = 11),
    panel.grid.minor = element_blank()
  )

# ══════════════════════════════════════════════════════════════════════════════
# 3. FIT K-MEANS  (change k here after inspecting elbow plot)
# ══════════════════════════════════════════════════════════════════════════════
k <- 3
km <- kmeans(df_scaled, centers = k, nstart = 25)

df <- df |>
  mutate(Cluster = factor(km$cluster, labels = paste("Cluster", 1:k)))

# ══════════════════════════════════════════════════════════════════════════════
# 4. SCATTER — Sleep Hours vs. each cognitive outcome, colored by cluster
# ══════════════════════════════════════════════════════════════════════════════
df_long <- df |>
  select(Sleep_Hours, Sleep_Quality_Score, Cluster, all_of(cognitive_vars)) |>
  pivot_longer(-c(Sleep_Hours, Sleep_Quality_Score, Cluster),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome = str_replace_all(outcome, "_", " "))

cluster_colors <- c("#4C72B0", "#C44E52", "#55A868", "#8172B2", "#CCB974")

p_scatter_hours <- ggplot(df_long,
                          aes(x = Sleep_Hours, y = value,
                              color = Cluster, fill = Cluster)) +
  geom_point(alpha = 0.4, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, alpha = 0.15) +
  scale_color_manual(values = cluster_colors) +
  scale_fill_manual(values  = cluster_colors) +
  facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
  labs(
    title    = "Sleep Hours vs. Cognitive Outcomes by Cluster",
    subtitle = "Clusters derived from k-means on cognitive performance profile",
    x        = "Sleep Hours", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

# ══════════════════════════════════════════════════════════════════════════════
# 5. CLUSTER PROFILE PLOT — what does each cluster actually look like?
# ══════════════════════════════════════════════════════════════════════════════
# Mean of all key variables per cluster, normalized for comparability
profile_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time", "Stress_Level"
)

cluster_profiles <- df |>
  select(Cluster, all_of(profile_vars)) |>
  group_by(Cluster) |>
  summarise(across(everything(), mean), .groups = "drop") |>
  pivot_longer(-Cluster, names_to = "variable", values_to = "mean_val") |>
  group_by(variable) |>
  mutate(scaled_mean = scale(mean_val)[,1]) |>  # z-score across clusters
  ungroup() |>
  mutate(variable = str_replace_all(variable, "_", " "))

p_profile <- ggplot(cluster_profiles,
                    aes(x = variable, y = scaled_mean,
                        color = Cluster, group = Cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  scale_color_manual(values = cluster_colors) +
  scale_x_discrete(guide = guide_axis(angle = 40)) +
  labs(
    title    = "Cluster Profiles — Mean Variable Values (Z-Scored)",
    subtitle = "Values above 0 = above average; below 0 = below average",
    x        = NULL, y        = "Z-Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

# ══════════════════════════════════════════════════════════════════════════════
# 6. BOXPLOTS — Sleep Hours per cluster (does sleep vary by cognitive cluster?)
# ══════════════════════════════════════════════════════════════════════════════
p_box <- df |>
  select(Cluster, Sleep_Hours, Sleep_Quality_Score) |>
  pivot_longer(-Cluster, names_to = "variable", values_to = "value") |>
  mutate(variable = str_replace_all(variable, "_", " ")) |>
  ggplot(aes(x = Cluster, y = value, fill = Cluster)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  scale_fill_manual(values = cluster_colors) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title    = "Sleep Distribution by Cognitive Cluster",
    subtitle = "Do clusters with better cognition sleep more or better?",
    x        = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "none"
  )

print(p_elbow)
print(p_scatter_hours)
print(p_profile)
print(p_box)

```

## Models

1\) Start with OLS as a baseline

```{r}

```

```{r}
library(tidyverse)

print_model_summary <- function(model, label) {
  cat("\n", strrep("=", 60), "\n")
  cat("MODEL:", label, "\n")
  cat(strrep("=", 60), "\n")
  print(summary(model))
  cat("\nVIF (Multicollinearity Check):\n")
  print(car::vif(model))
  cat("\n")
}

# =============================================================================
# PREDICTOR SETS
# =============================================================================
sleep_vars       <- c("Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness")
demo_vars        <- c("Age", "Gender", "BMI")
lifestyle_vars   <- c("Caffeine_Intake", "Physical_Activity_Level", "Stress_Level")

all_predictors   <- c(sleep_vars, demo_vars, lifestyle_vars)
predictor_formula <- paste(all_predictors, collapse = " + ")

# =============================================================================
# BLOCK 1: COGNITIVE PERFORMANCE OUTCOMES
# =============================================================================

# --- 1a. Stroop Task Reaction Time -------------------------------------------
# Higher = slower = worse cognitive inhibition
model_stroop <- lm(
  as.formula(paste("Stroop_Task_Reaction_Time ~", predictor_formula)),
  data = df
)
print_model_summary(model_stroop, "Stroop Task Reaction Time")

# --- 1b. N-Back Accuracy -----------------------------------------------------
# Higher = better working memory
model_nback <- lm(
  as.formula(paste("N_Back_Accuracy ~", predictor_formula)),
  data = df
)
print_model_summary(model_nback, "N-Back Accuracy")

# --- 1c. PVT Reaction Time ---------------------------------------------------
# Higher = slower = worse sustained attention
model_pvt <- lm(
  as.formula(paste("PVT_Reaction_Time ~", predictor_formula)),
  data = df
)
print_model_summary(model_pvt, "PVT Reaction Time")

# =============================================================================
# BLOCK 2: STRESS & EMOTION REGULATION OUTCOMES
# =============================================================================

# NOTE: Stress_Level is used as predictor for cognitive models above,
# but here we model it as an outcome — remove it from predictors to avoid
# circularity, and use sleep + demographics + activity instead.

stress_predictors <- c(sleep_vars, demo_vars, "Caffeine_Intake", "Physical_Activity_Level")
stress_formula    <- paste(stress_predictors, collapse = " + ")

# --- 2a. Stress Level --------------------------------------------------------
model_stress <- lm(
  as.formula(paste("Stress_Level ~", stress_formula)),
  data = df
)
print_model_summary(model_stress, "Stress Level")

# --- 2b. Emotion Regulation Score --------------------------------------------
model_emotion <- lm(
  as.formula(paste("Emotion_Regulation_Score ~", stress_formula)),
  data = df
)
print_model_summary(model_emotion, "Emotion Regulation Score")

# =============================================================================
# DIAGNOSTICS PLOTS (run per model)
# =============================================================================
plot_diagnostics <- function(model, label) {
  par(mfrow = c(2, 2))
  plot(model, main = label)
  par(mfrow = c(1, 1))
}

plot_diagnostics(model_stroop,  "Stroop")
plot_diagnostics(model_nback,   "N-Back")
plot_diagnostics(model_pvt,     "PVT")
plot_diagnostics(model_stress,  "Stress")
plot_diagnostics(model_emotion, "Emotion Regulation")

# =============================================================================
# MODEL COMPARISON TABLE (R², Adj R², F-stat, p-value)
# =============================================================================
model_list <- list(
  "Stroop RT"          = model_stroop,
  "N-Back Accuracy"    = model_nback,
  "PVT RT"             = model_pvt,
  "Stress Level"       = model_stress,
  "Emotion Regulation" = model_emotion
)

comparison_table <- map_dfr(model_list, function(m) {
  s <- summary(m)
  f <- s$fstatistic
  p_val <- pf(f[1], f[2], f[3], lower.tail = FALSE)
  tibble(
    R2        = round(s$r.squared, 3),
    Adj_R2    = round(s$adj.r.squared, 3),
    F_stat    = round(f[1], 3),
    df1       = f[2],
    df2       = f[3],
    p_value   = round(p_val, 4)
  )
}, .id = "Outcome")

cat("\n", strrep("=", 60), "\n")
cat("MODEL COMPARISON TABLE\n")
cat(strrep("=", 60), "\n")
print(comparison_table)

```

```{r}
# =============================================================================
# LASSO Regularization Models: Cognitive & Stress/Emotion Outcomes
# Uses cross-validated LASSO (glmnet) for automatic variable selection
# =============================================================================

library(tidyverse)
library(glmnet)

# --- Load Data ---------------------------------------------------------------
df <- df %>%
  mutate(Gender = as.numeric(factor(Gender)) - 1) %>%  # Binary encode: 0/1 for glmnet
  select(-Participant_ID)                               # Drop ID column — not a predictor

# =============================================================================
# HELPER FUNCTIONS
# =============================================================================

# Fit LASSO with cross-validation and return results
fit_lasso <- function(outcome_var, predictor_vars, data, label) {
  
  # Drop rows with NA in outcome or predictors
  model_data <- data %>%
    select(all_of(c(outcome_var, predictor_vars))) %>%
    drop_na()
  
  X <- model_data %>% select(all_of(predictor_vars)) %>% as.matrix()
  
  # Abort early with informative message if NAs remain
  if (any(is.na(X))) {
    na_cols <- colnames(X)[colSums(is.na(X)) > 0]
    stop("NAs found in predictor columns: ", paste(na_cols, collapse = ", "),
         "\nCheck your data for missing values or non-numeric columns.")
  }
  y <- model_data[[outcome_var]]
  
  # Scale predictors for fair penalization
  X_scaled <- scale(X)
  
  # 10-fold cross-validated LASSO (alpha = 1)
  set.seed(42)
  cv_fit <- cv.glmnet(X_scaled, y, alpha = 1, nfolds = 10)
  
  # Extract coefficients at lambda.min and lambda.1se
  coef_min  <- coef(cv_fit, s = "lambda.min")
  coef_1se  <- coef(cv_fit, s = "lambda.1se")
  
  # Selected variables (non-zero coefficients at lambda.1se — more parsimonious)
  selected_vars <- rownames(coef_1se)[which(coef_1se != 0)]
  selected_vars <- selected_vars[selected_vars != "(Intercept)"]
  
  # R² at lambda.min
  y_hat  <- predict(cv_fit, newx = X_scaled, s = "lambda.min")
  ss_res <- sum((y - y_hat)^2)
  ss_tot <- sum((y - mean(y))^2)
  r2     <- 1 - ss_res / ss_tot
  
  # Print results
  cat("\n", strrep("=", 60), "\n")
  cat("LASSO MODEL:", label, "\n")
  cat(strrep("=", 60), "\n")
  cat("Optimal lambda (min CV error):  ", round(cv_fit$lambda.min, 4), "\n")
  cat("Optimal lambda (1-SE rule):     ", round(cv_fit$lambda.1se, 4), "\n")
  cat("Cross-validated R² (lambda.min):", round(r2, 3), "\n\n")
  
  cat("--- Coefficients at lambda.1se (parsimonious model) ---\n")
  print(coef_1se[coef_1se[,1] != 0, , drop = FALSE])
  
  cat("\nSelected predictors:", 
      ifelse(length(selected_vars) > 0, 
             paste(selected_vars, collapse = ", "), 
             "None — all shrunk to zero"), "\n")
  
  # Return for further use
  list(
    cv_fit        = cv_fit,
    label         = label,
    outcome       = outcome_var,
    selected_vars = selected_vars,
    r2            = r2,
    lambda_min    = cv_fit$lambda.min,
    lambda_1se    = cv_fit$lambda.1se
  )
}

# Plot CV error curve
plot_lasso <- function(lasso_result) {
  plot(lasso_result$cv_fit,
       main = paste("CV Error —", lasso_result$label))
}

# =============================================================================
# PREDICTOR SETS (same logic as MLR script — avoid circularity)
# =============================================================================
cognitive_predictors <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Age", "Gender", "BMI",
  "Caffeine_Intake", "Physical_Activity_Level", "Stress_Level"
)

stress_predictors <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Age", "Gender", "BMI",
  "Caffeine_Intake", "Physical_Activity_Level"
)

# =============================================================================
# FIT ALL MODELS
# =============================================================================
lasso_stroop  <- fit_lasso("Stroop_Task_Reaction_Time", cognitive_predictors, df, "Stroop RT")
lasso_nback   <- fit_lasso("N_Back_Accuracy",           cognitive_predictors, df, "N-Back Accuracy")
lasso_pvt     <- fit_lasso("PVT_Reaction_Time",         cognitive_predictors, df, "PVT RT")
lasso_stress  <- fit_lasso("Stress_Level",              stress_predictors,    df, "Stress Level")
lasso_emotion <- fit_lasso("Emotion_Regulation_Score",  stress_predictors,    df, "Emotion Regulation")

# =============================================================================
# CV ERROR PLOTS (one per model)
# =============================================================================
par(mfrow = c(2, 3))
plot_lasso(lasso_stroop)
plot_lasso(lasso_nback)
plot_lasso(lasso_pvt)
plot_lasso(lasso_stress)
plot_lasso(lasso_emotion)
par(mfrow = c(1, 1))

# =============================================================================
# SUMMARY COMPARISON TABLE
# =============================================================================
all_models <- list(lasso_stroop, lasso_nback, lasso_pvt, lasso_stress, lasso_emotion)

summary_table <- map_dfr(all_models, function(m) {
  tibble(
    Outcome        = m$label,
    R2_lambda_min  = round(m$r2, 3),
    Lambda_min     = round(m$lambda_min, 4),
    Lambda_1se     = round(m$lambda_1se, 4),
    N_Selected     = length(m$selected_vars),
    Selected_Vars  = ifelse(length(m$selected_vars) > 0,
                            paste(m$selected_vars, collapse = ", "),
                            "None")
  )
})

cat("\n", strrep("=", 60), "\n")
cat("LASSO MODEL COMPARISON SUMMARY\n")
cat(strrep("=", 60), "\n")
print(summary_table, width = Inf)

# =============================================================================
# OPTIONAL: Refit OLS on LASSO-selected variables (for interpretable coefs)
# =============================================================================
refit_ols <- function(lasso_result, data) {
  vars <- lasso_result$selected_vars
  if (length(vars) == 0) {
    cat("\nNo variables selected for", lasso_result$label, "— skipping OLS refit.\n")
    return(NULL)
  }
  formula <- as.formula(paste(lasso_result$outcome, "~", paste(vars, collapse = " + ")))
  model   <- lm(formula, data = data)
  cat("\n--- OLS Refit on LASSO-selected vars:", lasso_result$label, "---\n")
  print(summary(model))
  model
}

ols_stroop  <- refit_ols(lasso_stroop,  df)
ols_nback   <- refit_ols(lasso_nback,   df)
ols_pvt     <- refit_ols(lasso_pvt,     df)
ols_stress  <- refit_ols(lasso_stress,  df)
ols_emotion <- refit_ols(lasso_emotion, df)
```
