---
title: "Sleep Deprivation and Cognitive Performance"
format: html
editor: visual
---

## Sleep Deprivation and Cognitive Performance

## Goals

-   Understand the effects of sleep deprivation (hours of sleep, sleep quality, daytime sleep) on cognitive performance (attention, working memory, reaction speed, emotion regulation)

### Read Data

```{r read_data}
library(readr)

df <- read_csv("sleep_deprivation_dataset_detailed.csv")
head(df)
```

### Data Structure

```{r data_structure}
str(df)
```

### Summary Statistics

```{r summary}
summary(df)
```

```{r unique_gender}
cat("Unique values of df$Gender:", unique(df$Gender), "\n")
cat("Unique values of df$Caffeine_Intake", sort(unique(df$Caffeine_Intake)), "\n")
```

Note:

-   `Caffeine_Intake`, `Sleep_Quality_Score`, `Daytime_Sleepiness`, `Physical_Activity_Level`, and `Stress_Level` are discrete variables.

-   `Gender` is a binary variable.

### Univariate Distrubutions

```{r continuous_dists}
library(ggplot2)
library(tidyverse)
library(dplyr)

continuous_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time",
  "Age", "BMI", "Stress_Level"
)

df_long_cont <- df |>
  select(all_of(continuous_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(variable = str_replace_all(variable, "_", " "))

p_continuous <- ggplot(df_long_cont, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "#4C72B0", color = "white", alpha = 0.8) +
  geom_density(color = "#C44E52", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(
    title    = "Univariate Distributions — Continuous Variables",
    subtitle = "Histogram with density overlay",
    x = NULL, y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 7)
  )

p_continuous
```

```{r discrete_dists}
library(ggplot2)
library(tidyverse)
library(dplyr)

discrete_long <- bind_rows(
  df |> count(value = factor(Caffeine_Intake))       |> mutate(variable = "Caffeine Intake (1–5)"),
  df |> count(value = factor(Physical_Activity_Level)) |> mutate(variable = "Physical Activity Level (1–10)"),
  df |> count(value = factor(Gender))                |> mutate(variable = "Gender")
)

p_discrete <- ggplot(discrete_long, aes(x = value, y = n, fill = variable)) +
  geom_col(show.legend = FALSE, width = 0.6, alpha = 0.85) +
  geom_text(aes(label = n), vjust = -0.4, size = 3) +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title    = "Univariate Distributions — Discrete & Categorical Variables",
    subtitle = "Count per category",
    x = NULL, y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title        = element_text(face = "bold", size = 14),
    plot.subtitle     = element_text(color = "grey50", size = 11),
    strip.text        = element_text(face = "bold", size = 9),
    panel.grid.major.x = element_blank(),
    panel.grid.minor  = element_blank()
  )

p_discrete
```

### Correlation Matrix

```{r corrplot}
library(tidyverse)
library(ggcorrplot)

numeric_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time",
  "Age", "BMI", "Caffeine_Intake", "Physical_Activity_Level", "Stress_Level"
)

cor_long <- df |>
  select(all_of(numeric_vars)) |>
  rename_with(~ str_replace_all(.x, "_", "\n")) |>
  cor(use = "pairwise.complete.obs") |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  # keep lower triangle only
  mutate(var1 = factor(var1, levels = unique(var1)),
         var2 = factor(var2, levels = unique(var2))) |>
  filter(as.integer(var1) > as.integer(var2))

p_corr <- ggplot(cor_long, aes(x = var2, y = var1, size = abs(r), fill = r)) +
  geom_point(shape = 21, color = "white") +
  geom_text(aes(label = round(r, 2)), size = 3, color = "grey20") +
  scale_fill_gradient2(low = "#C44E52", mid = "white", high = "#4C72B0",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  scale_size_continuous(range = c(2, 14), guide = "none") +
  labs(
    title    = "Correlation Matrix",
    subtitle = "Circle size = |r|",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "grey50", size = 11),
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7, lineheight = 0.85),
    axis.text.y     = element_text(size = 7, lineheight = 0.85),
    panel.grid      = element_blank(),
    legend.position = "right",
    plot.margin     = margin(10, 10, 10, 30)
  )

p_corr
```

### Scatterplots

```{r scatter_sleep_vs_cognition}
library(tidyverse)
library(ggplot2)
library(dplyr)

plot_sleep_vs_cognition <- function(data, sleep_var, sleep_label) {
  
  cognitive_vars <- c(
    "Stroop_Task_Reaction_Time",
    "N_Back_Accuracy",
    "Emotion_Regulation_Score",
    "PVT_Reaction_Time"
  )
  
  data |>
    select(all_of(c(sleep_var, cognitive_vars))) |>
    pivot_longer(-all_of(sleep_var), names_to = "outcome", values_to = "value") |>
    mutate(outcome = str_replace_all(outcome, "_", " ")) |>
    ggplot(aes(x = .data[[sleep_var]], y = value)) +
    geom_point(alpha = 0.4, size = 1.8, color = "#4C72B0") +
    geom_smooth(method = "lm", se = TRUE, color = "#C44E52", linewidth = 1) +
    facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
    labs(
      title    = paste("Sleep vs. Cognitive Outcomes —", sleep_label),
      subtitle = "Linear fit with 95% confidence interval",
      x        = sleep_label,
      y        = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title     = element_text(face = "bold", size = 15),
      plot.subtitle  = element_text(color = "grey50", size = 11),
      strip.text     = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank()
    )
}

# ── Generate both plots ───────────────────────────────────────────────────────
p_hours   <- plot_sleep_vs_cognition(df, "Sleep_Hours",         "Sleep Hours")
p_quality <- plot_sleep_vs_cognition(df, "Sleep_Quality_Score", "Sleep Quality Score")

print(p_hours)
print(p_quality)
```

### Binned Scatterplots

```{r binned_scatterplots}
library(tidyverse)
library(ggplot2)
library(dplyr)

df <- df |>
  mutate(Age_Group = cut(Age,
                         breaks = c(-Inf, 25, 35, 45, Inf),
                         labels = c("≤25", "26–35", "36–45", "46+")))

plot_binned <- function(data, sleep_var, sleep_label, group_var, group_label) {
  
  cognitive_vars <- c(
    "Stroop_Task_Reaction_Time",
    "N_Back_Accuracy",
    "Emotion_Regulation_Score",
    "PVT_Reaction_Time"
  )
  
  data |>
    select(all_of(c(sleep_var, group_var, cognitive_vars))) |>
    pivot_longer(-all_of(c(sleep_var, group_var)),
                 names_to = "outcome", values_to = "value") |>
    mutate(outcome = str_replace_all(outcome, "_", " ")) |>
    ggplot(aes(x = .data[[sleep_var]], y = value,
               color = .data[[group_var]], fill = .data[[group_var]])) +
    geom_point(alpha = 0.3, size = 1.5) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 1, alpha = 0.15) +
    facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
    labs(
      title    = paste("Sleep vs. Cognitive Outcomes —", sleep_label),
      subtitle = paste("By", group_label, "| Linear fit with 95% CI"),
      x        = sleep_label,
      y        = NULL,
      color    = group_label,
      fill     = group_label
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey50", size = 11),
      strip.text       = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
}


p_hours_gender   <- plot_binned(df, "Sleep_Hours",         "Sleep Hours",         "Gender", "Gender")
p_quality_gender <- plot_binned(df, "Sleep_Quality_Score", "Sleep Quality Score", "Gender", "Gender")

p_hours_age   <- plot_binned(df, "Sleep_Hours",         "Sleep Hours",         "Age_Group", "Age Group")
p_quality_age <- plot_binned(df, "Sleep_Quality_Score", "Sleep Quality Score", "Age_Group", "Age Group")


gender_colors <- c("#4C72B0", "#C44E52", "#55A868")  # flexible for 2–3 gender categories
age_colors    <- c("#4C72B0", "#55A868", "#C44E52", "#8172B2")

p_hours_gender   <- p_hours_gender   + scale_color_manual(values = gender_colors) + scale_fill_manual(values = gender_colors)
p_quality_gender <- p_quality_gender + scale_color_manual(values = gender_colors) + scale_fill_manual(values = gender_colors)
p_hours_age      <- p_hours_age      + scale_color_manual(values = age_colors)    + scale_fill_manual(values = age_colors)
p_quality_age    <- p_quality_age    + scale_color_manual(values = age_colors)    + scale_fill_manual(values = age_colors)

print(p_hours_gender)
print(p_quality_gender)
print(p_hours_age)
print(p_quality_age)
```
