---
title: "Sleep Deprivation and Cognitive Performance"
format: html
editor: visual
---

## Sleep Deprivation and Cognitive Performance

## Goals

-   Understand the effects of sleep deprivation (hours of sleep, sleep quality, daytime sleep) on cognitive performance (attention, working memory, reaction speed, emotion regulation)
-   Model cognitive performance against sleep quality, naps, and caffeine intake

### Read Data

```{r read_data}
library(readr)

df <- read_csv("sleep_deprivation_dataset_detailed.csv")
head(df)
```

### Data Structure

```{r data_structure}
str(df)
```

### Summary Statistics

```{r summary}
summary(df)
```

```{r unique_gender}
cat("Unique values of df$Gender:", unique(df$Gender), "\n")
cat("Unique values of df$Caffeine_Intake", sort(unique(df$Caffeine_Intake)), "\n")
```

Note:

-   `Caffeine_Intake`, `Sleep_Quality_Score`, `Daytime_Sleepiness`, `Physical_Activity_Level`, and `Stress_Level` are discrete variables.

-   `Gender` is a binary variable.

### Univariate Distrubutions

```{r continuous_dists}
library(ggplot2)
library(tidyverse)
library(dplyr)

continuous_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time",
  "Age", "BMI", "Stress_Level"
)

df_long_cont <- df |>
  select(all_of(continuous_vars)) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  mutate(variable = str_replace_all(variable, "_", " "))

p_continuous <- ggplot(df_long_cont, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30, fill = "#4C72B0", color = "white", alpha = 0.8) +
  geom_density(color = "#C44E52", linewidth = 0.8) +
  facet_wrap(~ variable, scales = "free", ncol = 4) +
  labs(
    title    = "Univariate Distributions — Continuous Variables",
    subtitle = "Histogram with density overlay",
    x = NULL, y = "Density"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title       = element_text(face = "bold", size = 14),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 9),
    panel.grid.minor = element_blank(),
    axis.text.x      = element_text(size = 7)
  )

p_continuous
```

```{r discrete_dists}
library(ggplot2)
library(tidyverse)
library(dplyr)

discrete_long <- bind_rows(
  df |> count(value = factor(Caffeine_Intake))       |> mutate(variable = "Caffeine Intake (1–5)"),
  df |> count(value = factor(Physical_Activity_Level)) |> mutate(variable = "Physical Activity Level (1–10)"),
  df |> count(value = factor(Gender))                |> mutate(variable = "Gender")
)

p_discrete <- ggplot(discrete_long, aes(x = value, y = n, fill = variable)) +
  geom_col(show.legend = FALSE, width = 0.6, alpha = 0.85) +
  geom_text(aes(label = n), vjust = -0.4, size = 3) +
  scale_fill_brewer(palette = "Set2") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  labs(
    title    = "Univariate Distributions — Discrete & Categorical Variables",
    subtitle = "Count per category",
    x = NULL, y = "Count"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title        = element_text(face = "bold", size = 14),
    plot.subtitle     = element_text(color = "grey50", size = 11),
    strip.text        = element_text(face = "bold", size = 9),
    panel.grid.major.x = element_blank(),
    panel.grid.minor  = element_blank()
  )

p_discrete
```

### Correlation Matrix

```{r corrplot}
library(tidyverse)
library(ggcorrplot)

numeric_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time",
  "Age", "BMI", "Caffeine_Intake", "Physical_Activity_Level", "Stress_Level"
)

cor_long <- df |>
  select(all_of(numeric_vars)) |>
  rename_with(~ str_replace_all(.x, "_", "\n")) |>
  cor(use = "pairwise.complete.obs") |>
  as.data.frame() |>
  rownames_to_column("var1") |>
  pivot_longer(-var1, names_to = "var2", values_to = "r") |>
  # keep lower triangle only
  mutate(var1 = factor(var1, levels = unique(var1)),
         var2 = factor(var2, levels = unique(var2))) |>
  filter(as.integer(var1) > as.integer(var2))

p_corr <- ggplot(cor_long, aes(x = var2, y = var1, size = abs(r), fill = r)) +
  geom_point(shape = 21, color = "white") +
  geom_text(aes(label = round(r, 2)), size = 3, color = "grey20") +
  scale_fill_gradient2(low = "#C44E52", mid = "white", high = "#4C72B0",
                       midpoint = 0, limits = c(-1, 1), name = "r") +
  scale_size_continuous(range = c(2, 14), guide = "none") +
  labs(
    title    = "Correlation Matrix",
    subtitle = "Circle size = |r|",
    x = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold", size = 15),
    plot.subtitle   = element_text(color = "grey50", size = 11),
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 7, lineheight = 0.85),
    axis.text.y     = element_text(size = 7, lineheight = 0.85),
    panel.grid      = element_blank(),
    legend.position = "right",
    plot.margin     = margin(10, 10, 10, 30)
  )

p_corr
```

### Scatterplots

```{r scatter_sleep_vs_cognition}
library(tidyverse)
library(ggplot2)
library(dplyr)

plot_sleep_vs_cognition <- function(data, sleep_var, sleep_label) {
  
  cognitive_vars <- c(
    "Stroop_Task_Reaction_Time",
    "N_Back_Accuracy",
    "Emotion_Regulation_Score",
    "PVT_Reaction_Time"
  )
  
  data |>
    select(all_of(c(sleep_var, cognitive_vars))) |>
    pivot_longer(-all_of(sleep_var), names_to = "outcome", values_to = "value") |>
    mutate(outcome = str_replace_all(outcome, "_", " ")) |>
    ggplot(aes(x = .data[[sleep_var]], y = value)) +
    geom_point(alpha = 0.4, size = 1.8, color = "#4C72B0") +
    geom_smooth(method = "lm", se = TRUE, color = "#C44E52", linewidth = 1) +
    facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
    labs(
      title    = paste("Sleep vs. Cognitive Outcomes —", sleep_label),
      subtitle = "Linear fit with 95% confidence interval",
      x        = sleep_label,
      y        = NULL
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title     = element_text(face = "bold", size = 15),
      plot.subtitle  = element_text(color = "grey50", size = 11),
      strip.text     = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank()
    )
}

# ── Generate both plots ───────────────────────────────────────────────────────
p_hours   <- plot_sleep_vs_cognition(df, "Sleep_Hours",         "Sleep Hours")
p_quality <- plot_sleep_vs_cognition(df, "Sleep_Quality_Score", "Sleep Quality Score")

print(p_hours)
print(p_quality)
```

### Binned Scatterplots

```{r binned_scatterplots}
library(tidyverse)
library(ggplot2)
library(dplyr)

df <- df |>
  mutate(Age_Group = cut(Age,
                         breaks = c(-Inf, 25, 35, 45, Inf),
                         labels = c("≤25", "26–35", "36–45", "46+")))

plot_binned <- function(data, sleep_var, sleep_label, group_var, group_label) {
  
  cognitive_vars <- c(
    "Stroop_Task_Reaction_Time",
    "N_Back_Accuracy",
    "Emotion_Regulation_Score",
    "PVT_Reaction_Time"
  )
  
  data |>
    select(all_of(c(sleep_var, group_var, cognitive_vars))) |>
    pivot_longer(-all_of(c(sleep_var, group_var)),
                 names_to = "outcome", values_to = "value") |>
    mutate(outcome = str_replace_all(outcome, "_", " ")) |>
    ggplot(aes(x = .data[[sleep_var]], y = value,
               color = .data[[group_var]], fill = .data[[group_var]])) +
    geom_point(alpha = 0.3, size = 1.5) +
    geom_smooth(method = "lm", se = TRUE, linewidth = 1, alpha = 0.15) +
    facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
    labs(
      title    = paste("Sleep vs. Cognitive Outcomes —", sleep_label),
      subtitle = paste("By", group_label, "| Linear fit with 95% CI"),
      x        = sleep_label,
      y        = NULL,
      color    = group_label,
      fill     = group_label
    ) +
    theme_minimal(base_size = 13) +
    theme(
      plot.title       = element_text(face = "bold", size = 15),
      plot.subtitle    = element_text(color = "grey50", size = 11),
      strip.text       = element_text(face = "bold", size = 10),
      panel.grid.minor = element_blank(),
      legend.position  = "bottom"
    )
}


p_hours_gender   <- plot_binned(df, "Sleep_Hours",         "Sleep Hours",         "Gender", "Gender")
p_quality_gender <- plot_binned(df, "Sleep_Quality_Score", "Sleep Quality Score", "Gender", "Gender")

p_hours_age   <- plot_binned(df, "Sleep_Hours",         "Sleep Hours",         "Age_Group", "Age Group")
p_quality_age <- plot_binned(df, "Sleep_Quality_Score", "Sleep Quality Score", "Age_Group", "Age Group")


gender_colors <- c("#4C72B0", "#C44E52", "#55A868")  # flexible for 2–3 gender categories
age_colors    <- c("#4C72B0", "#55A868", "#C44E52", "#8172B2")

p_hours_gender   <- p_hours_gender   + scale_color_manual(values = gender_colors) + scale_fill_manual(values = gender_colors)
p_quality_gender <- p_quality_gender + scale_color_manual(values = gender_colors) + scale_fill_manual(values = gender_colors)
p_hours_age      <- p_hours_age      + scale_color_manual(values = age_colors)    + scale_fill_manual(values = age_colors)
p_quality_age    <- p_quality_age    + scale_color_manual(values = age_colors)    + scale_fill_manual(values = age_colors)

print(p_hours_gender)
print(p_quality_gender)
print(p_hours_age)
print(p_quality_age)
```

### Clustering

```{r cluster_by_sleep_group}
library(tidyverse)
library(dplyr)

df <- df |>
  mutate(Sleep_Group = cut(Sleep_Hours,
                           breaks = c(-Inf, 5, 6, 7, 9, Inf),
                           labels = c("Minimal\n(<5h)", "Poor\n(5–6h)",
                                      "Decent\n(6–7h)", "Sufficient\n(7–9h)",
                                      "Excessive\n(>9h)")))

# ── Pivot cognitive outcomes to long format ───────────────────────────────────
cognitive_vars <- c(
  "Stroop_Task_Reaction_Time",
  "N_Back_Accuracy",
  "Emotion_Regulation_Score",
  "PVT_Reaction_Time"
)

df_long <- df |>
  select(Sleep_Hours, Sleep_Group, all_of(cognitive_vars)) |>
  pivot_longer(-c(Sleep_Hours, Sleep_Group),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome = str_replace_all(outcome, "_", " "))

# ── Color palette: intuitive red → green spectrum ────────────────────────────
sleep_colors <- c(
  "Minimal\n(<5h)"    = "#C44E52",
  "Poor\n(5–6h)"      = "#DD8452",
  "Decent\n(6–7h)"    = "#CCB974",
  "Sufficient\n(7–9h)"= "#55A868",
  "Excessive\n(>9h)"  = "#4C72B0"
)

# ── Plot ──────────────────────────────────────────────────────────────────────
p <- ggplot(df_long, aes(x = Sleep_Hours, y = value,
                         color = Sleep_Group, fill = Sleep_Group)) +
  geom_point(alpha = 0.4, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, alpha = 0.15) +
  scale_color_manual(values = sleep_colors, name = "Sleep Duration") +
  scale_fill_manual(values  = sleep_colors, name = "Sleep Duration") +
  facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
  labs(
    title    = "Sleep Hours vs. Cognitive Outcomes by Sleep Duration Group",
    subtitle = "Each group has its own linear fit | Points colored by sleep duration bin",
    x        = "Sleep Hours",
    y        = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom",
    legend.title     = element_text(face = "bold"),
    legend.text      = element_text(size = 9)
  ) +
  guides(color = guide_legend(nrow = 1, override.aes = list(alpha = 1, size = 3)),
         fill  = guide_legend(nrow = 1))

p
```

```{r}
cognitive_vars <- c(
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score",  "PVT_Reaction_Time"
)

df_scaled <- df |>
  select(all_of(cognitive_vars)) |>
  scale() |>
  as.data.frame()

# ══════════════════════════════════════════════════════════════════════════════
# 2. ELBOW PLOT — choose optimal k
# ══════════════════════════════════════════════════════════════════════════════
set.seed(42)
wss <- map_dbl(1:8, ~ kmeans(df_scaled, centers = .x, nstart = 25)$tot.withinss)

p_elbow <- tibble(k = 1:8, wss = wss) |>
  ggplot(aes(x = k, y = wss)) +
  geom_line(linewidth = 1, color = "grey50") +
  geom_point(size = 4, color = "#4C72B0") +
  scale_x_continuous(breaks = 1:8) +
  labs(
    title    = "Elbow Plot — K-Means on Cognitive Outcomes",
    subtitle = "Look for the 'elbow' where WSS stops dropping sharply",
    x        = "Number of Clusters (k)",
    y        = "Total Within-Cluster SS"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(color = "grey50", size = 11),
    panel.grid.minor = element_blank()
  )

# ══════════════════════════════════════════════════════════════════════════════
# 3. FIT K-MEANS  (change k here after inspecting elbow plot)
# ══════════════════════════════════════════════════════════════════════════════
k <- 3
km <- kmeans(df_scaled, centers = k, nstart = 25)

df <- df |>
  mutate(Cluster = factor(km$cluster, labels = paste("Cluster", 1:k)))

# ══════════════════════════════════════════════════════════════════════════════
# 4. SCATTER — Sleep Hours vs. each cognitive outcome, colored by cluster
# ══════════════════════════════════════════════════════════════════════════════
df_long <- df |>
  select(Sleep_Hours, Sleep_Quality_Score, Cluster, all_of(cognitive_vars)) |>
  pivot_longer(-c(Sleep_Hours, Sleep_Quality_Score, Cluster),
               names_to = "outcome", values_to = "value") |>
  mutate(outcome = str_replace_all(outcome, "_", " "))

cluster_colors <- c("#4C72B0", "#C44E52", "#55A868", "#8172B2", "#CCB974")

p_scatter_hours <- ggplot(df_long,
                          aes(x = Sleep_Hours, y = value,
                              color = Cluster, fill = Cluster)) +
  geom_point(alpha = 0.4, size = 1.8) +
  geom_smooth(method = "lm", se = TRUE, linewidth = 1, alpha = 0.15) +
  scale_color_manual(values = cluster_colors) +
  scale_fill_manual(values  = cluster_colors) +
  facet_wrap(~ outcome, scales = "free_y", ncol = 2) +
  labs(
    title    = "Sleep Hours vs. Cognitive Outcomes by Cluster",
    subtitle = "Clusters derived from k-means on cognitive performance profile",
    x        = "Sleep Hours", y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

# ══════════════════════════════════════════════════════════════════════════════
# 5. CLUSTER PROFILE PLOT — what does each cluster actually look like?
# ══════════════════════════════════════════════════════════════════════════════
# Mean of all key variables per cluster, normalized for comparability
profile_vars <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Stroop_Task_Reaction_Time", "N_Back_Accuracy",
  "Emotion_Regulation_Score", "PVT_Reaction_Time", "Stress_Level"
)

cluster_profiles <- df |>
  select(Cluster, all_of(profile_vars)) |>
  group_by(Cluster) |>
  summarise(across(everything(), mean), .groups = "drop") |>
  pivot_longer(-Cluster, names_to = "variable", values_to = "mean_val") |>
  group_by(variable) |>
  mutate(scaled_mean = scale(mean_val)[,1]) |>  # z-score across clusters
  ungroup() |>
  mutate(variable = str_replace_all(variable, "_", " "))

p_profile <- ggplot(cluster_profiles,
                    aes(x = variable, y = scaled_mean,
                        color = Cluster, group = Cluster)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey70") +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3.5) +
  scale_color_manual(values = cluster_colors) +
  scale_x_discrete(guide = guide_axis(angle = 40)) +
  labs(
    title    = "Cluster Profiles — Mean Variable Values (Z-Scored)",
    subtitle = "Values above 0 = above average; below 0 = below average",
    x        = NULL, y        = "Z-Score"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

# ══════════════════════════════════════════════════════════════════════════════
# 6. BOXPLOTS — Sleep Hours per cluster (does sleep vary by cognitive cluster?)
# ══════════════════════════════════════════════════════════════════════════════
p_box <- df |>
  select(Cluster, Sleep_Hours, Sleep_Quality_Score) |>
  pivot_longer(-Cluster, names_to = "variable", values_to = "value") |>
  mutate(variable = str_replace_all(variable, "_", " ")) |>
  ggplot(aes(x = Cluster, y = value, fill = Cluster)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.4) +
  scale_fill_manual(values = cluster_colors) +
  facet_wrap(~ variable, scales = "free_y") +
  labs(
    title    = "Sleep Distribution by Cognitive Cluster",
    subtitle = "Do clusters with better cognition sleep more or better?",
    x        = NULL, y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title       = element_text(face = "bold", size = 15),
    plot.subtitle    = element_text(color = "grey50", size = 11),
    strip.text       = element_text(face = "bold", size = 10),
    panel.grid.minor = element_blank(),
    legend.position  = "none"
  )

print(p_elbow)
print(p_scatter_hours)
print(p_profile)
print(p_box)

```

## Models

```{r}
# =============================================================================
# Cognitive Performance Models: Stepwise Selection + LASSO Regularization
# Outcomes:  Stroop_Task_Reaction_Time, N_Back_Accuracy, PVT_Reaction_Time
# Predictors: Sleep (Hours, Quality, Sleepiness), Caffeine, Age, Gender, BMI
# =============================================================================

library(tidyverse)
library(glmnet)
library(MASS)  # for stepAIC

# NOTE: Assumes df is already loaded with Gender as factor

# =============================================================================
# SETUP
# =============================================================================
outcomes <- c("Stroop_Task_Reaction_Time", "N_Back_Accuracy", "PVT_Reaction_Time")

predictors <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Caffeine_Intake",
  "Age", "Gender", "BMI"
)

# Clean working copy — only keep needed columns, drop NA rows
model_df <- df %>%
  dplyr::select(all_of(c(outcomes, predictors))) %>%
  drop_na()

cat("Rows used for modeling:", nrow(model_df), "\n")

# =============================================================================
# METHOD 1: STEPWISE SELECTION (AIC-based, bidirectional)
# =============================================================================
run_stepwise <- function(outcome, data, predictors) {
  full_formula <- as.formula(paste(outcome, "~", paste(predictors, collapse = " + ")))
  null_formula <- as.formula(paste(outcome, "~ 1"))

  full_model <- lm(full_formula, data = data)
  null_model <- lm(null_formula, data = data)

  # Bidirectional stepwise (both forward and backward)
  step_model <- stepAIC(
    null_model,
    scope = list(lower = null_formula, upper = full_formula),
    direction = "both",
    trace = FALSE
  )

  cat("\n", strrep("=", 60), "\n")
  cat("STEPWISE —", outcome, "\n")
  cat(strrep("=", 60), "\n")
  cat("Selected formula: "); print(formula(step_model))
  print(summary(step_model))

  step_model
}

step_stroop <- run_stepwise("Stroop_Task_Reaction_Time", model_df, predictors)
step_nback  <- run_stepwise("N_Back_Accuracy",           model_df, predictors)
step_pvt    <- run_stepwise("PVT_Reaction_Time",         model_df, predictors)

# =============================================================================
# METHOD 2: LASSO REGULARIZATION (cv.glmnet, alpha = 1)
# =============================================================================

# glmnet needs a numeric matrix — encode Gender as 0/1
model_df_num <- model_df %>%
  mutate(Gender = as.numeric(factor(Gender)) - 1)

run_lasso <- function(outcome, data, predictors) {
  X <- data %>% dplyr::select(all_of(predictors)) %>% as.matrix()
  y <- data[[outcome]]
  X_scaled <- scale(X)

  set.seed(42)
  cv_fit <- cv.glmnet(X_scaled, y, alpha = 1, nfolds = 10)

  coef_1se     <- coef(cv_fit, s = "lambda.1se")
  selected     <- rownames(coef_1se)[coef_1se[, 1] != 0]
  selected     <- selected[selected != "(Intercept)"]

  y_hat  <- as.vector(predict(cv_fit, newx = X_scaled, s = "lambda.min"))
  r2     <- 1 - sum((y - y_hat)^2) / sum((y - mean(y))^2)

  cat("\n", strrep("=", 60), "\n")
  cat("LASSO —", outcome, "\n")
  cat(strrep("=", 60), "\n")
  cat("lambda.min:", round(cv_fit$lambda.min, 4),
      " | lambda.1se:", round(cv_fit$lambda.1se, 4), "\n")
  cat("R² at lambda.min:", round(r2, 3), "\n")
  cat("Selected predictors (lambda.1se):",
      ifelse(length(selected) > 0, paste(selected, collapse = ", "), "None"), "\n")
  cat("\nCoefficients at lambda.1se:\n")
  print(coef_1se[coef_1se[, 1] != 0, , drop = FALSE])

  list(cv_fit = cv_fit, selected = selected, r2 = r2, outcome = outcome)
}

lasso_stroop <- run_lasso("Stroop_Task_Reaction_Time", model_df_num, predictors)
lasso_nback  <- run_lasso("N_Back_Accuracy",           model_df_num, predictors)
lasso_pvt    <- run_lasso("PVT_Reaction_Time",         model_df_num, predictors)

# =============================================================================
# LASSO CV PLOTS
# =============================================================================
par(mfrow = c(1, 3))
plot(lasso_stroop$cv_fit, main = "LASSO — Stroop RT")
plot(lasso_nback$cv_fit,  main = "LASSO — N-Back Accuracy")
plot(lasso_pvt$cv_fit,    main = "LASSO — PVT RT")
par(mfrow = c(1, 1))

# =============================================================================
# LASSO → OLS REFIT on selected variables (unbiased coefs for reporting)
# =============================================================================
refit_ols <- function(lasso_result, data) {
  vars    <- lasso_result$selected
  outcome <- lasso_result$outcome

  if (length(vars) == 0) {
    cat("\nNo variables selected by LASSO for", outcome, "— skipping OLS refit.\n")
    return(NULL)
  }

  formula <- as.formula(paste(outcome, "~", paste(vars, collapse = " + ")))
  ols     <- lm(formula, data = data)

  cat("\n--- OLS Refit on LASSO-selected vars:", outcome, "---\n")
  print(summary(ols))
  ols
}

ols_stroop <- refit_ols(lasso_stroop, model_df)
ols_nback  <- refit_ols(lasso_nback,  model_df)
ols_pvt    <- refit_ols(lasso_pvt,    model_df)

# =============================================================================
# SUMMARY COMPARISON TABLE
# =============================================================================
extract_step_stats <- function(model, label, method) {
  s   <- summary(model)
  f   <- s$fstatistic
  if (is.null(f)) return(tibble(Method=method, Outcome=label, R2=NA, Adj_R2=NA, p_value=NA, Selected_Vars="(intercept only)"))
  tibble(
    Method        = method,
    Outcome       = label,
    R2            = round(s$r.squared, 3),
    Adj_R2        = round(s$adj.r.squared, 3),
    p_value       = round(pf(f[1], f[2], f[3], lower.tail = FALSE), 4),
    Selected_Vars = paste(names(coef(model))[-1], collapse = ", ")
  )
}

extract_lasso_stats <- function(lasso_result, label) {
  tibble(
    Method        = "LASSO",
    Outcome       = label,
    R2            = round(lasso_result$r2, 3),
    Adj_R2        = NA,
    p_value       = NA,
    Selected_Vars = ifelse(length(lasso_result$selected) > 0,
                           paste(lasso_result$selected, collapse = ", "), "None")
  )
}

comparison <- bind_rows(
  extract_step_stats(step_stroop,  "Stroop RT",       "Stepwise"),
  extract_step_stats(step_nback,   "N-Back Accuracy", "Stepwise"),
  extract_step_stats(step_pvt,     "PVT RT",          "Stepwise"),
  extract_lasso_stats(lasso_stroop, "Stroop RT"),
  extract_lasso_stats(lasso_nback,  "N-Back Accuracy"),
  extract_lasso_stats(lasso_pvt,    "PVT RT")
) %>% arrange(Outcome, Method)

cat("\n", strrep("=", 60), "\n")
cat("FINAL COMPARISON TABLE\n")
cat(strrep("=", 60), "\n")
print(comparison, width = Inf)
```

```{r}
library(tidyverse)

print_model_summary <- function(model, label) {
  cat("\n", strrep("=", 60), "\n")
  cat("MODEL:", label, "\n")
  cat(strrep("=", 60), "\n")
  print(summary(model))
  cat("\nVIF (Multicollinearity Check):\n")
  print(car::vif(model))
  cat("\n")
}

# =============================================================================
# PREDICTOR SETS
# =============================================================================
sleep_vars       <- c("Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness")
demo_vars        <- c("Age", "Gender", "BMI")
lifestyle_vars   <- c("Caffeine_Intake", "Physical_Activity_Level", "Stress_Level")

all_predictors   <- c(sleep_vars, demo_vars, lifestyle_vars)
predictor_formula <- paste(all_predictors, collapse = " + ")

# =============================================================================
# BLOCK 1: COGNITIVE PERFORMANCE OUTCOMES
# =============================================================================

# --- 1a. Stroop Task Reaction Time -------------------------------------------
# Higher = slower = worse cognitive inhibition
model_stroop <- lm(
  as.formula(paste("Stroop_Task_Reaction_Time ~", predictor_formula)),
  data = df
)
print_model_summary(model_stroop, "Stroop Task Reaction Time")

# --- 1b. N-Back Accuracy -----------------------------------------------------
# Higher = better working memory
model_nback <- lm(
  as.formula(paste("N_Back_Accuracy ~", predictor_formula)),
  data = df
)
print_model_summary(model_nback, "N-Back Accuracy")

# --- 1c. PVT Reaction Time ---------------------------------------------------
# Higher = slower = worse sustained attention
model_pvt <- lm(
  as.formula(paste("PVT_Reaction_Time ~", predictor_formula)),
  data = df
)
print_model_summary(model_pvt, "PVT Reaction Time")

# =============================================================================
# BLOCK 2: STRESS & EMOTION REGULATION OUTCOMES
# =============================================================================

# NOTE: Stress_Level is used as predictor for cognitive models above,
# but here we model it as an outcome — remove it from predictors to avoid
# circularity, and use sleep + demographics + activity instead.

stress_predictors <- c(sleep_vars, demo_vars, "Caffeine_Intake", "Physical_Activity_Level")
stress_formula    <- paste(stress_predictors, collapse = " + ")

# --- 2a. Stress Level --------------------------------------------------------
model_stress <- lm(
  as.formula(paste("Stress_Level ~", stress_formula)),
  data = df
)
print_model_summary(model_stress, "Stress Level")

# --- 2b. Emotion Regulation Score --------------------------------------------
model_emotion <- lm(
  as.formula(paste("Emotion_Regulation_Score ~", stress_formula)),
  data = df
)
print_model_summary(model_emotion, "Emotion Regulation Score")

# =============================================================================
# DIAGNOSTICS PLOTS (run per model)
# =============================================================================
plot_diagnostics <- function(model, label) {
  par(mfrow = c(2, 2))
  plot(model, main = label)
  par(mfrow = c(1, 1))
}

plot_diagnostics(model_stroop,  "Stroop")
plot_diagnostics(model_nback,   "N-Back")
plot_diagnostics(model_pvt,     "PVT")
plot_diagnostics(model_stress,  "Stress")
plot_diagnostics(model_emotion, "Emotion Regulation")

# =============================================================================
# MODEL COMPARISON TABLE (R², Adj R², F-stat, p-value)
# =============================================================================
model_list <- list(
  "Stroop RT"          = model_stroop,
  "N-Back Accuracy"    = model_nback,
  "PVT RT"             = model_pvt,
  "Stress Level"       = model_stress,
  "Emotion Regulation" = model_emotion
)

comparison_table <- map_dfr(model_list, function(m) {
  s <- summary(m)
  f <- s$fstatistic
  p_val <- pf(f[1], f[2], f[3], lower.tail = FALSE)
  tibble(
    R2        = round(s$r.squared, 3),
    Adj_R2    = round(s$adj.r.squared, 3),
    F_stat    = round(f[1], 3),
    df1       = f[2],
    df2       = f[3],
    p_value   = round(p_val, 4)
  )
}, .id = "Outcome")

cat("\n", strrep("=", 60), "\n")
cat("MODEL COMPARISON TABLE\n")
cat(strrep("=", 60), "\n")
print(comparison_table)

```

```{r}
# =============================================================================
# LASSO Regularization Models: Cognitive & Stress/Emotion Outcomes
# Uses cross-validated LASSO (glmnet) for automatic variable selection
# =============================================================================

library(tidyverse)
library(glmnet)

# --- Load Data ---------------------------------------------------------------
df <- df %>%
  mutate(Gender = as.numeric(factor(Gender)) - 1) %>%  # Binary encode: 0/1 for glmnet
  select(-Participant_ID)                               # Drop ID column — not a predictor

# =============================================================================
# HELPER FUNCTIONS
# =============================================================================

# Fit LASSO with cross-validation and return results
fit_lasso <- function(outcome_var, predictor_vars, data, label) {
  
  # Drop rows with NA in outcome or predictors
  model_data <- data %>%
    select(all_of(c(outcome_var, predictor_vars))) %>%
    drop_na()
  
  X <- model_data %>% select(all_of(predictor_vars)) %>% as.matrix()
  
  # Abort early with informative message if NAs remain
  if (any(is.na(X))) {
    na_cols <- colnames(X)[colSums(is.na(X)) > 0]
    stop("NAs found in predictor columns: ", paste(na_cols, collapse = ", "),
         "\nCheck your data for missing values or non-numeric columns.")
  }
  y <- model_data[[outcome_var]]
  
  # Scale predictors for fair penalization
  X_scaled <- scale(X)
  
  # 10-fold cross-validated LASSO (alpha = 1)
  set.seed(42)
  cv_fit <- cv.glmnet(X_scaled, y, alpha = 1, nfolds = 10)
  
  # Extract coefficients at lambda.min and lambda.1se
  coef_min  <- coef(cv_fit, s = "lambda.min")
  coef_1se  <- coef(cv_fit, s = "lambda.1se")
  
  # Selected variables (non-zero coefficients at lambda.1se — more parsimonious)
  selected_vars <- rownames(coef_1se)[which(coef_1se != 0)]
  selected_vars <- selected_vars[selected_vars != "(Intercept)"]
  
  # R² at lambda.min
  y_hat  <- predict(cv_fit, newx = X_scaled, s = "lambda.min")
  ss_res <- sum((y - y_hat)^2)
  ss_tot <- sum((y - mean(y))^2)
  r2     <- 1 - ss_res / ss_tot
  
  # Print results
  cat("\n", strrep("=", 60), "\n")
  cat("LASSO MODEL:", label, "\n")
  cat(strrep("=", 60), "\n")
  cat("Optimal lambda (min CV error):  ", round(cv_fit$lambda.min, 4), "\n")
  cat("Optimal lambda (1-SE rule):     ", round(cv_fit$lambda.1se, 4), "\n")
  cat("Cross-validated R² (lambda.min):", round(r2, 3), "\n\n")
  
  cat("--- Coefficients at lambda.1se (parsimonious model) ---\n")
  print(coef_1se[coef_1se[,1] != 0, , drop = FALSE])
  
  cat("\nSelected predictors:", 
      ifelse(length(selected_vars) > 0, 
             paste(selected_vars, collapse = ", "), 
             "None — all shrunk to zero"), "\n")
  
  # Return for further use
  list(
    cv_fit        = cv_fit,
    label         = label,
    outcome       = outcome_var,
    selected_vars = selected_vars,
    r2            = r2,
    lambda_min    = cv_fit$lambda.min,
    lambda_1se    = cv_fit$lambda.1se
  )
}

# Plot CV error curve
plot_lasso <- function(lasso_result) {
  plot(lasso_result$cv_fit,
       main = paste("CV Error —", lasso_result$label))
}

# =============================================================================
# PREDICTOR SETS (same logic as MLR script — avoid circularity)
# =============================================================================
cognitive_predictors <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Age", "Gender", "BMI",
  "Caffeine_Intake", "Physical_Activity_Level", "Stress_Level"
)

stress_predictors <- c(
  "Sleep_Hours", "Sleep_Quality_Score", "Daytime_Sleepiness",
  "Age", "Gender", "BMI",
  "Caffeine_Intake", "Physical_Activity_Level"
)

# =============================================================================
# FIT ALL MODELS
# =============================================================================
lasso_stroop  <- fit_lasso("Stroop_Task_Reaction_Time", cognitive_predictors, df, "Stroop RT")
lasso_nback   <- fit_lasso("N_Back_Accuracy",           cognitive_predictors, df, "N-Back Accuracy")
lasso_pvt     <- fit_lasso("PVT_Reaction_Time",         cognitive_predictors, df, "PVT RT")
lasso_stress  <- fit_lasso("Stress_Level",              stress_predictors,    df, "Stress Level")
lasso_emotion <- fit_lasso("Emotion_Regulation_Score",  stress_predictors,    df, "Emotion Regulation")

# =============================================================================
# CV ERROR PLOTS (one per model)
# =============================================================================
par(mfrow = c(2, 3))
plot_lasso(lasso_stroop)
plot_lasso(lasso_nback)
plot_lasso(lasso_pvt)
plot_lasso(lasso_stress)
plot_lasso(lasso_emotion)
par(mfrow = c(1, 1))

# =============================================================================
# SUMMARY COMPARISON TABLE
# =============================================================================
all_models <- list(lasso_stroop, lasso_nback, lasso_pvt, lasso_stress, lasso_emotion)

summary_table <- map_dfr(all_models, function(m) {
  tibble(
    Outcome        = m$label,
    R2_lambda_min  = round(m$r2, 3),
    Lambda_min     = round(m$lambda_min, 4),
    Lambda_1se     = round(m$lambda_1se, 4),
    N_Selected     = length(m$selected_vars),
    Selected_Vars  = ifelse(length(m$selected_vars) > 0,
                            paste(m$selected_vars, collapse = ", "),
                            "None")
  )
})

cat("\n", strrep("=", 60), "\n")
cat("LASSO MODEL COMPARISON SUMMARY\n")
cat(strrep("=", 60), "\n")
print(summary_table, width = Inf)

# =============================================================================
# OPTIONAL: Refit OLS on LASSO-selected variables (for interpretable coefs)
# =============================================================================
refit_ols <- function(lasso_result, data) {
  vars <- lasso_result$selected_vars
  if (length(vars) == 0) {
    cat("\nNo variables selected for", lasso_result$label, "— skipping OLS refit.\n")
    return(NULL)
  }
  formula <- as.formula(paste(lasso_result$outcome, "~", paste(vars, collapse = " + ")))
  model   <- lm(formula, data = data)
  cat("\n--- OLS Refit on LASSO-selected vars:", lasso_result$label, "---\n")
  print(summary(model))
  model
}

ols_stroop  <- refit_ols(lasso_stroop,  df)
ols_nback   <- refit_ols(lasso_nback,   df)
ols_pvt     <- refit_ols(lasso_pvt,     df)
ols_stress  <- refit_ols(lasso_stress,  df)
ols_emotion <- refit_ols(lasso_emotion, df)
```
